---
title: "MPhil Thesis: HT Week 2 Exploration"
author: "Mohammad-Shah Karim"
date: "26 Jan 2025"

execute:
  dir: project

format:
  revealjs:
    css: styles.css
    toc: true
    toc-depth: 1
    slideNumber: true
    auto: true
    scrollable: true
    embed-resources: true
    self-contained-math: true
---

```{r}
library(here)

#exploration/HT_week1_RQ1.r


# Load necessary libraries
library(tidyverse)
library(lubridate)
library(readr)
library(stringr)
library(ggplot2)
library(fixest)
library(knitr)
library(modelsummary)
library(kableExtra)
library(fixest)
library(TwoWayFEWeights)
library(DIDmultiplegtDYN)
library(did)
library(here)


# Load FDA approval data

fda_approvals <- read_csv(
    here(
        "processed_data/fda_approvals/fda_sdud_pdl_merged.csv"
        )
) |>
rename(
    approvals = n_approvals,
    firm = sponsor_name
    )

# Load patent data

patents <- read_csv(
    here(
        "processed_data/patents/patents_sdud_pdl_merged.csv"
    )
) |>
    rename(
        atc_level1 = ATC1,
        patents = num_patents)

# --- Build ex-ante demand --- #

# FDA

fda_ex_ante_stats <- fda_approvals |>
    filter(approval_year <= 2013 & approval_year >= 2010) |>
    distinct(atc_level1, approval_year, approval_quarter, prescriptions)

fda_ex_ante_total_prescriptions <- fda_ex_ante_stats |>
    summarise(
        total_prescriptions = sum(prescriptions, na.rm = TRUE),
        .groups = "drop"
    ) |>
    pull()

fda_ex_ante_demand <- fda_ex_ante_stats |>
    group_by(atc_level1) |>
    summarise(
        d_ex_ante = sum(prescriptions, na.rm = TRUE) / fda_ex_ante_total_prescriptions,
        .groups = "drop"
    )

# Patents

patents_ex_ante_stats <- patents |>
    filter(patent_year <= 2013 & patent_year >= 2010) |>
    distinct(atc_level1, patent_year, patent_quarter, prescriptions)

patents_ex_ante_total_prescriptions <- patents_ex_ante_stats |>
    summarise(
        total_prescriptions = sum(prescriptions, na.rm = TRUE),
        .groups = "drop"
    ) |>
    pull()

patents_ex_ante_demand <- patents_ex_ante_stats |>
    group_by(atc_level1) |>
    summarise(
        d_ex_ante = sum(prescriptions, na.rm = TRUE) / patents_ex_ante_total_prescriptions,
        .groups = "drop"
    )


# --- Build ex-post demand --- #

# FDA

fda_ex_post_stats <- fda_approvals |>
    filter(approval_year %in% 2014:2015) |>
    distinct(atc_level1, approval_year, approval_quarter, prescriptions)

fda_ex_post_total_prescriptions <- fda_ex_post_stats |>
    summarise(
        total_prescriptions = sum(prescriptions, na.rm = TRUE),
        .groups = "drop"
    ) |>
    pull()

fda_ex_post_demand <- fda_ex_post_stats |>
    group_by(atc_level1) |>
    summarise(
        d_ex_post = sum(prescriptions, na.rm = TRUE) / fda_ex_post_total_prescriptions,
        .groups = "drop"
    )

# Patents

patents_ex_post_stats <- patents |>
    filter(patent_year %in% 2014:2015) |>
    distinct(atc_level1, patent_year, patent_quarter, prescriptions)

patents_ex_post_total_prescriptions <- patents_ex_post_stats |>
    summarise(
        total_prescriptions = sum(prescriptions, na.rm = TRUE),
        .groups = "drop"
    ) |>
    pull()

patents_ex_post_demand <- patents_ex_post_stats |>
    group_by(atc_level1) |>
    summarise(
        d_ex_post = sum(prescriptions, na.rm = TRUE) / patents_ex_post_total_prescriptions,
        .groups = "drop"
    )


# --- Merge demand measures --- #

fda_approvals <- fda_approvals |>
  left_join(fda_ex_ante_demand, by = "atc_level1") |>
  left_join(fda_ex_post_demand, by = "atc_level1") |>
  filter(approval_year >= 2010)

patents <- patents |>
  left_join(patents_ex_ante_demand, by = "atc_level1") |>
  left_join(patents_ex_post_demand, by = "atc_level1") |>
  filter(patent_year >= 2010)


# --- Run regressions for FDA data --- #

fda_ex_ante_data <- fda_approvals |>
  select(-d_ex_post) |>
  rename(medicaid_market_share = d_ex_ante)

fda_ex_post_data <- fda_approvals |>
  select(-d_ex_ante) |>
  rename(medicaid_market_share = d_ex_post)

patents_ex_ante_data <- patents |>
  select(-d_ex_post) |>
  rename(medicaid_market_share = d_ex_ante)

patents_ex_post_data <- patents |>
  select(-d_ex_ante) |>
  rename(medicaid_market_share = d_ex_post)


```

# Research Question 1: Overall innovation response to Market size

## The model

$$
\begin{aligned}
Y_{f,c,t}
= \alpha_{f, c} + \delta_t
+ \sum_{k \neq 2013}^{2025} \beta_k \left[ \mathbf{1}\{t = k\} \times \text{MedicaidShare}_c \right]
+ \varepsilon_{f,c,t}
\end{aligned}
$$

## RQ1: FDA plots

```{r}

fda_rq1_feols_exante <- feols(
  approvals ~ i(approval_year, medicaid_market_share, ref = 2013) |
              firm^atc_level1 + approval_year,
              data = fda_ex_ante_data,
              cluster = ~ firm
)

iplot(
  fda_rq1_feols_exante,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "FEOLS FDA approvals ex-ante exposure"
  )

rm(fda_rq1_feols_exante)

fda_rq1_feols_expost <- feols(
  approvals ~ i(approval_year, medicaid_market_share, ref = 2013) |
              firm^atc_level1 + approval_year,
              data = fda_ex_post_data,
              cluster = ~ firm
)

iplot(
  fda_rq1_feols_expost,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "FEOLS FDA approvals ex-post exposure"
  )

rm(fda_rq1_feols_expost)

fda_rq1_poi_exante <- fepois(
  approvals ~ i(approval_year, medicaid_market_share, ref = 2013) |
              firm^atc_level1 + approval_year,
              data = fda_ex_ante_data,
              cluster = ~ firm
)

iplot(
  fda_rq1_poi_exante,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "FE Poisson approvals ex-ante exposure"
  )

rm(fda_rq1_poi_exante)

fda_rq1_poi_expost <- fepois(
  approvals ~ i(approval_year, medicaid_market_share, ref = 2013) |
              firm^atc_level1 + approval_year,
              data = fda_ex_post_data,
              cluster = ~ firm
)

iplot(
  fda_rq1_poi_expost,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "FE Poisson approvals ex-post exposure"
  )

rm(fda_rq1_poi_expost)

```

## RQ1: Patent plots

```{r}

patent_rq1_feols_exante <- feols(
  patents ~ i(patent_year, medicaid_market_share, ref = 2013) |
              firm^atc_level1 + patent_year,
              data = patents_ex_ante_data,
              cluster = ~ firm
)

iplot(
  patent_rq1_feols_exante,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "FEOLS patents ex-ante exposure"
  )

rm(patent_rq1_feols_exante)

patent_rq1_feols_expost <- feols(
  patents ~ i(patent_year, medicaid_market_share, ref = 2013) |
              firm^atc_level1 + patent_year,
              data = patents_ex_post_data,
              cluster = ~ firm
)

iplot(
  patent_rq1_feols_expost,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "FEOLS patents ex-post exposure"
  )

rm(patent_rq1_feols_expost)

patent_rq1_poi_exante <- fepois(
  patents ~ i(patent_year, medicaid_market_share, ref = 2013) |
              firm^atc_level1 + patent_year,
              data = patents_ex_ante_data,
              cluster = ~ firm
)

iplot(
  patent_rq1_poi_exante,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "NB patents ex-ante exposure"
  )

rm(patent_rq1_poi_exante)

patent_rq1_poi_expost <- fepois(
  patents ~ i(patent_year, medicaid_market_share, ref = 2013) |
              firm^atc_level1 + patent_year,
              data = patents_ex_post_data,
              cluster = ~ firm
)

iplot(
  patent_rq1_poi_expost,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "NB patents ex-post exposure"
  )

rm(patent_rq1_poi_expost)

```

# Research Question 2: Market Size and Pricing Pressure

## The model

$$
\begin{aligned}
Y_{f,c,t}
&= \alpha_{f,c} + \delta_t + \sum_{l \neq 2013}^{2025} \beta_{l}
\left( \text{Expansion}_{l} \times \text{MedicaidShare}_c \right) \\
&\quad + \sum_{j \neq 2013}^{2025} \gamma_{\ell}
\left( \text{Expansion}_{j} \times \text{PDL_Exposure}_{f,c} \right) + \varepsilon_{f,c,t}
\end{aligned}
$$

Where:

$$
\text{PDL_Exposure}_{f,c}
=
\frac{
\text{Number of drugs on PDL}_{f,c}
}{
\text{Number of drugs}_{f,c}
}
$$

- $\beta$ measures market size effects
- $\gamma$ measures the effect of firms more exposed to the PDL pricing pressure

- Measuring price effects is unrealistic and therefore measuring the effect of being on the market itself is worhtwhl

## FDA plots

```{r}

fda_rq2_feols_exante <- feols(
  approvals ~ i(approval_year, medicaid_market_share, ref = 2013) 
            + i(approval_year, in_pdl, ref = 2013) | 
              firm^atc_level1 + approval_year,
              data = fda_ex_ante_data,
              cluster = ~ firm
)

iplot(
  fda_rq2_feols_exante,
  i.select = 1,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "Event study: Medicaid market size (ex ante)"
)

iplot(
  fda_rq2_feols_exante,
  i.select = 2,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "Event study: PDL exposure (ex ante)"
)

rm(fda_rq2_feols_exante)

fda_rq2_feols_expost <- feols(
  approvals ~ i(approval_year, medicaid_market_share, ref = 2013) 
            + i(approval_year, in_pdl, ref = 2013) |
              firm^atc_level1 + approval_year,
              data = fda_ex_post_data,
              cluster = ~ firm
)

iplot(
  fda_rq2_feols_expost,
  i.select = 1,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "Event study: Medicaid market size (ex post)"
)

iplot(
  fda_rq2_feols_expost,
  i.select = 2,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "Event study: PDL exposure (ex post)"
)

rm(fda_rq2_feols_expost)

```

## Patent plots

```{r}


patents_ex_ante_year <- patents_ex_ante_data |>
  group_by(firm, atc_level1, patent_year) |>
  summarise(
    patents = sum(patents, na.rm = TRUE),
    medicaid_market_share = first(medicaid_market_share),
    in_pdl = max(in_pdl, na.rm = TRUE),
    .groups = "drop"
  )


patents_ex_post_year <- patents_ex_post_data |>
  group_by(firm, atc_level1, patent_year) |>
  summarise(
    patents = sum(patents, na.rm = TRUE),
    medicaid_market_share = first(medicaid_market_share),
    in_pdl = max(in_pdl, na.rm = TRUE),
    .groups = "drop"
  )



patent_rq2_feols_exante <- feols(
  patents ~ i(patent_year, medicaid_market_share, ref = 2013)
            + i(patent_year, in_pdl, ref = 2013) | 
              firm^atc_level1 + patent_year,
              data = patents_ex_ante_year,
              cluster = ~ firm
)

iplot(
  patent_rq2_feols_exante,
  i.select = 1,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "Event study: Medicaid market size (ex ante)"
)

iplot(
  patent_rq2_feols_exante,
  i.select = 2,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "Event study: PDL exposure (ex ante)"
)

rm(patent_rq2_feols_exante)

patent_rq2_feols_expost <- feols(
  patents ~ i(patent_year, medicaid_market_share, ref = 2013)
            + i(patent_year, in_pdl, ref = 2013) |  
              firm^atc_level1 + patent_year,
              data = patents_ex_post_year,
              cluster = ~ firm
)

iplot(
  patent_rq2_feols_expost,
  i.select = 1,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "Event study: Medicaid market size (ex post)"
)

iplot(
  patent_rq2_feols_expost,
  i.select = 2,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "Event study: PDL exposure (ex post)"
)

rm(patent_rq2_feols_expost)

```

# Research Question 3: Innovation portfolio

## The model

$$
\begin{aligned}
\begin{aligned}
Y_{f,c,t}
&= \alpha_f + \lambda_c + \delta_t \\
&\quad + \sum_{c \in \mathcal{C}}
\beta_c
\Big(
\mathbb{1}\{t \geq 2014\}
\times \text{MedicaidShare}_c
\times \mathbb{1}\{\text{ATC} = c\}
\Big)
+ \varepsilon_{f,c,t}.
\end{aligned}
\end{aligned}
$$

## FDA coefficients

```{r}

fda_ex_ante_data <- fda_ex_ante_data |>
  mutate(post = ifelse(approval_year >= 2014, 1, 0))

fda_ex_post_data <- fda_ex_post_data |>
  mutate(post = ifelse(approval_year >= 2014, 1, 0))

fda_rq3_feols_exante <- feols(
  approvals ~ medicaid_market_share : post : i(atc_level1) | 
  firm + atc_level1 + approval_year,
  data = fda_ex_ante_data,
  cluster = ~ firm
)

fda_rq3_feols_expost <- feols(
  approvals ~ medicaid_market_share : post : i(atc_level1) | 
  firm + atc_level1 + approval_year,
  data = fda_ex_post_data,
  cluster = ~ firm
)

fda_rq3_model_list <- list(
  "FEOLS: Ex Ante"  = fda_rq3_feols_exante,
  "FEOLS: Ex Post"  = fda_rq3_feols_expost
)

fda_rq3_models <- modelsummary(
             fda_rq3_model_list,
             stars = c(`*` = 0.05, `**` = 0.01, `***` = 0.001)
             )

fda_rq3_models

rm(fda_rq3_feols_exante, fda_rq3_feols_expost, fda_rq3_model_list, fda_rq3_models)

rm(fda_approvals)

```

## Patent coefficients

```{r}

patents_ex_ante_year <- patents_ex_ante_year |>
  mutate(post = ifelse(patent_year >= 2014, 1, 0))

patents_ex_post_year <- patents_ex_post_year |>
  mutate(post = ifelse(patent_year >= 2014, 1, 0))

patent_rq3_feols_exante <- feols(
  patents ~ medicaid_market_share : post : i(atc_level1) | 
  firm + atc_level1 + patent_year,
  data = patents_ex_ante_year,
  cluster = ~ firm
)

patent_rq3_feols_expost <- feols(
  patents ~ medicaid_market_share : post : i(atc_level1) | 
  firm + atc_level1 + patent_year,
  data = patents_ex_post_year,
  cluster = ~ firm
)


patent_rq3_model_list <- list(
  "FEOLS: Ex Ante"  = patent_rq3_feols_exante,
  "FEOLS: Ex Post"  = patent_rq3_feols_expost
)

patent_rq3_models <- modelsummary(
             patent_rq3_model_list,
             stars = c(`*` = 0.05, `**` = 0.01, `***` = 0.001)
             )


patent_rq3_models

rm(patent_rq3_feols_exante, patent_rq3_feols_expost, patent_rq3_model_list, patent_rq3_models)

```