---
title: "PDL Counterfactual Analysis"
author: "Mohammad-Shah Karim"
date: today

execute:
  dir: project

format:
  revealjs:
    css: styles.css
    toc: true
    toc-depth: 1
    slideNumber: true
    auto: true
    scrollable: true
    embed-resources: true
    self-contained-math: true
    mermaid:
      theme: default
---

# The Preferred Drug List

## The Mechanism

A drug becomes preferred when it is clinically acceptable **and** offers the lowest net cost among substitutes.

1. P&T committee evaluates safety, efficacy and literature on drug.

2. States compare net price among clinically equivalent drugs, with price determining which eligible drugs become most preferred.

```{mermaid}
flowchart LR
    Start["Candidate Drug"]
    
    Clinical["Clinical Review<br/>(Safety & Efficacy)"]
    Economic["Economic Evaluation<br/>(Net Cost vs Substitutes)"]
    
    Start --> Clinical
    Start --> Economic

    Clinical --> Decision{"Composite<br/>Decision"}
    Economic --> Decision

    Decision -->|PASSED BOTH<br/>(Clinical + Cost)| PDL["Preferred Drug<br/>(Winner)"]
    Decision -->|FAILED EITHER<br/>(Clinical or Cost)| NonPDL["Non-Preferred Drug<br/>(Counterfactuals)"]
```


## Understanding counterfactuals

There are two situations by which a pharmaceutical firm gets rejected:

1. **Failing the P&T stage**: The drug does not reach standards for efficacy or safety.

2. **Failing the cost stage**: Some other drug exists that is cheaper for Medicaid to purchase for that state.

In both cases, failing only one of these stages indicates that a firm had just missed out on being eligible for the Medicaid PDL.

**However**

It's likely that incentives differ based on which stage had failed.

# Patent data

```{r}

# Load necessary libraries
library(tidyverse)
library(lubridate)
library(readr)
library(stringr)
library(ggplot2)
library(fixest)
library(knitr)
library(modelsummary)
library(kableExtra)
library(fixest)
library(TwoWayFEWeights)
library(DIDmultiplegtDYN)
library(did)
library(here)

# Load FDA approval data

fda_approvals <- read_csv(
    here(
        "processed_data/fda_approvals/fda_sdud_pdl_merged.csv"
        )
) |>
rename(
    approvals = n_approvals,
    firm = sponsor_name
    )

# Load patent data

patents <- read_csv(
    here(
        "processed_data/patents/patents_sdud_pdl_merged.csv"
    )
) |>
    rename(
        atc_level1 = ATC1,
        patents = num_patents) |>
    filter(patent_year >= 2010)

```

## P&T pass distribution

```{r}
patents_pt <- patents %>% filter(pt_stage_pass > 0)

ggplot(patents_pt, aes(x = pt_stage_pass)) +
  geom_histogram(
    bins = 30,
    fill = "steelblue",
    color = "white"
  ) +
  labs(
    title = "Distribution of P&T pass rates",
    x = "P&T pass rate (share passing stage)",
    y = "Count"
  ) +
  theme_classic()

```


## Price stage pass distribution

```{r}

patents_price <- patents %>% filter(price_stage_pass > 0)

ggplot(patents_price, aes(x = price_stage_pass)) +
  geom_histogram(
    bins = 30,
    fill = "steelblue",
    color = "white"
  ) +
  labs(
    title = "Distribution of Price pass rates",
    x = "Price pass rate (share passing stage)",
    y = "Count"
  ) +
  theme_classic()

```

## In PDL distribution

```{r}

patents_pdl <- patents %>% filter(in_pdl > 0)

ggplot(patents_pdl, aes(x = in_pdl)) +
  geom_histogram(
    bins = 30,
    fill = "steelblue",
    color = "white"
  ) +
  labs(
    title = "Distribution in PDL",
    x = "Share in PDL",
    y = "Count"
  ) +
  theme_classic()

```


# Empirical Modelling

## The effect of passing the price stage

- Only using observations with $P\&T = 1$.
- This means the counterfactual is those that do not fail the price stage test.

$$
Y_{f,c,t} = \alpha_{f,c} + \delta_t + \beta \mathbb{1}\{Year_t \geq 2014\} \times \text{In_PDL}_{f,c} + \phi 
$$



## The effect of passing the P&T Stage

- Only using observations with $\text{Price_Stage} = 1$.
- This means the counterfactual is those that do not fail P&T stage.

$$
Y_{f,c,t} = \alpha_{f,c} + \delta_t + \beta \mathbb{1}\{Year_t \geq 2014\} \times \text{In_PDL}_{f,c}
$$

## Passing first stage and passing both

$$
Y_{f,c,t} = \alpha_{f,c} + \delta_t + \beta \mathbb{1}\{Year_t \geq 2014\} \times \text{In_PDL}_{f,c}
$$

# Empirical Results

## Passing the price stage test

```r

patents_price_stage <- patents |>
    filter(pt_stage_pass == 1) |>
    mutate(post = ifelse(patent_year >= 2014, 1, 0))


pass_price_stage_test <- feols(
  patents ~ in_pdl : post | 
              firm^atc_level1 + patent_year,
              data = patents_price_stage,
              cluster = ~ firm
)

modelsummary(pass_price_stage_test)

```

```{r}

patents_price_stage <- patents |>
    filter(pt_stage_pass == 1) |>
    mutate(post = ifelse(patent_year >= 2014, 1, 0))


pass_price_stage_test <- feols(
  patents ~ in_pdl : post | 
              firm^atc_level1 + patent_year,
              data = patents_price_stage,
              cluster = ~ firm
)

modelsummary(pass_price_stage_test,
             stars = c(`*` = 0.05, `**` = 0.01, `***` = 0.001),
             gof_omit = "R2 Within*|R2 Adj*|RMSE|Std.Errors|IC|Log",)

```

## Passing the P&T stage

```r

patents_pt_stage <- patents |>
    filter(price_stage_pass == 1) |>
    mutate(post = ifelse(patent_year >= 2014, 1, 0))


pass_pt_stage_test <- feols(
  patents ~ in_pdl : post | 
              firm^atc_level1 + patent_year,
              data = patents_pt_stage,
              cluster = ~ firm
)

modelsummary(pass_pt_stage_test)

```

```{r}

patents_pt_stage <- patents |>
    filter(price_stage_pass == 1) |>
    mutate(post = ifelse(patent_year >= 2014, 1, 0))


pass_pt_stage_test <- feols(
  patents ~ in_pdl : post | 
              firm^atc_level1 + patent_year,
              data = patents_pt_stage,
              cluster = ~ firm
)

modelsummary(pass_pt_stage_test,
             stars = c(`*` = 0.05, `**` = 0.01, `***` = 0.001),
             gof_omit = "R2 Within*|R2 Adj*|RMSE|Std.Errors|IC|Log",)

```


## Passing first stage and passing both

```{r}

patents <- patents |>
    mutate(post = ifelse(patent_year >= 2014, 1, 0))


overall_pdl_test <- feols(
  patents ~ post : pt_stage_pass + price_stage_pass : post + in_pdl : post | 
              firm^atc_level1 + patent_year,
              data = patents,
              cluster = ~ firm
)

modelsummary(overall_pdl_test,
             stars = c(`*` = 0.05, `**` = 0.01, `***` = 0.001),
             gof_omit = "R2 Within*|R2 Adj*|RMSE|Std.Errors|IC|Log",)

pt_stage_coef <- as.numeric(coefficients(overall_pdl_test)[1])
price_stage_coef <- as.numeric(coefficients(overall_pdl_test)[2])
in_pdl_coef <- as.numeric(coefficients(overall_pdl_test)[3])

```

## The "Stranded Quality" Penalty


| Regulatory Signal | Coefficient | Economic Interpretation |
| --- | --- | --- |
| **Clinical Pass Only**<br>($\beta_{PT}$) | **`r round(pt_stage_coef, 3)`** | **The "Stranded Quality" Penalty.**<br>Firms with high clinical quality ($a_i \uparrow$) that fail the price check face the steepest decline. |
| **Price Pass Only**<br>($\beta_{Price}$) | **`r round(price_stage_coef, 3)`** | **The "Low-Quality" Penalty.**<br>Low-cost drugs that fail clinical standards are also discouraged. |
| **PDL Winner**<br>($\beta_{PDL}$) | **+`r round(in_pdl_coef, 3)`** | **The "Access Premium."**<br>Securing the contract reverses the penalty. The market rewards access, not just science. |

**Conclusion:**
The aggregate decline in innovation is driven by **Unrewarded Quality**.
The expansion didn't make *all* innovation bad; it specifically punished expensive innovation that could not clear the stricter budget hurdle ($\gamma \uparrow$).