---
title: "PDL Counterfactual Analysis"
author: "Mohammad-Shah Karim"
date: today

execute:
  dir: project

format:
  revealjs:
    css: styles.css
    toc: true
    toc-depth: 1
    slideNumber: true
    auto: true
    scrollable: true
    embed-resources: true
    self-contained-math: true
    mermaid:
      theme: default
---

# The Preferred Drug List

## The Mechanism

A drug becomes preferred when it is clinically acceptable **and** offers the lowest net cost among substitutes.

1. P&T committee evaluates safety, efficacy and literature on drug.

2. States compare net price among clinically equivalent drugs, with price determining which eligible drugs become most preferred.

```{mermaid}
flowchart LR
    Start["Candidate Drug"]
    
    Clinical["Clinical Review<br>(Safety & Efficacy)"]
    Economic["Economic Evaluation<br>(Net Cost vs Substitutes)"]
    
    Start --> Clinical
    Start --> Economic

    Clinical --> Decision{"Composite<br>Decision"}
    Economic --> Decision

    Decision -->|"PASSED BOTH<br>(Clinical + Cost)"| PDL["Preferred Drug<br>(Winner)"]
    Decision -->|"FAILED EITHER<br>(Clinical or Cost)"| NonPDL["Non-Preferred Drug<br>(Counterfactuals)"]
```


## Understanding counterfactuals

There are two situations by which a pharmaceutical firm gets rejected:

1. **Failing the P&T stage**: The drug does not reach standards for efficacy or safety.

2. **Failing the cost stage**: Some other drug exists that is cheaper for Medicaid to purchase for that state.

In both cases, failing only one of these stages indicates that a firm had just missed out on being eligible for the Medicaid PDL.

**However**

It's likely that incentives differ based on which stage had failed.

# Patent data

```{r}

# Load necessary libraries
library(tidyverse)
library(lubridate)
library(readr)
library(stringr)
library(ggplot2)
library(fixest)
library(knitr)
library(modelsummary)
library(kableExtra)
library(fixest)
library(TwoWayFEWeights)
library(DIDmultiplegtDYN)
library(did)
library(here)

# Load FDA approval data

fda_approvals <- read_csv(
    here(
        "processed_data/fda_approvals/fda_sdud_pdl_merged.csv"
        )
) |>
rename(
    approvals = n_approvals,
    firm = sponsor_name
    )

# Load patent data

patents <- read_csv(
    here(
        "processed_data/patents/patents_sdud_pdl_merged.csv"
    )
) |>
    rename(
        atc_level1 = ATC1,
        patents = num_patents) |>
    filter(patent_year >= 2010)

```

## P&T pass distribution

```{r}
patents_pt <- patents %>% filter(pt_stage_pass > 0)

ggplot(patents_pt, aes(x = pt_stage_pass)) +
  geom_histogram(
    bins = 30,
    fill = "steelblue",
    color = "white"
  ) +
  labs(
    title = "Distribution of P&T pass rates",
    x = "P&T pass rate (share passing stage)",
    y = "Count"
  ) +
  theme_classic()

```


## Price stage pass distribution

```{r}

patents_price <- patents %>% filter(price_stage_pass > 0)

ggplot(patents_price, aes(x = price_stage_pass)) +
  geom_histogram(
    bins = 30,
    fill = "steelblue",
    color = "white"
  ) +
  labs(
    title = "Distribution of Price pass rates",
    x = "Price pass rate (share passing stage)",
    y = "Count"
  ) +
  theme_classic()

```

## In PDL distribution

```{r}

patents_pdl <- patents %>% filter(in_pdl > 0)

ggplot(patents_pdl, aes(x = in_pdl)) +
  geom_histogram(
    bins = 30,
    fill = "steelblue",
    color = "white"
  ) +
  labs(
    title = "Distribution in PDL",
    x = "Share in PDL",
    y = "Count"
  ) +
  theme_classic()

```


# Passing first stage and passing both

## Empirical modelling

$$
\begin{aligned}
Y_{f,c,t}
&= \alpha_{f,c} + \delta_t \\
&\quad + \beta_1 \big( PT\_Pass \times Post \big) \\
&\quad + \beta_2 \big( Price\_Pass \times Post \big) \\
&\quad + \beta_3 \big( \underbrace{PT\_Pass \times Price\_Pass}_{\text{In PDL}}
\times Post \big)
+ \varepsilon_{f,c,t}
\end{aligned}
$$

- Since $In\_PDL \approx \text{Clinical Pass} \times \text{Price Pass}$, the model is effectively a **Triple Difference** specification.
- This isolates the value of passing each stage *independently* versus passing *both*.

## Empirical results

```{r}

patents <- patents |>
    mutate(post = ifelse(patent_year >= 2014, 1, 0))


overall_pdl_test <- feols(
  patents ~ post : pt_stage_pass + price_stage_pass : post + in_pdl : post | 
              firm^atc_level1 + patent_year,
              data = patents,
              cluster = ~ firm
)

modelsummary(overall_pdl_test,
             stars = c(`*` = 0.05, `**` = 0.01, `***` = 0.001),
             gof_omit = "R2 Within*|R2 Adj*|RMSE|Std.Errors|IC|Log")

pt_stage_coef <- as.numeric(coefficients(overall_pdl_test)[1])
price_stage_coef <- as.numeric(coefficients(overall_pdl_test)[2])
in_pdl_coef <- as.numeric(coefficients(overall_pdl_test)[3])

```

## Interpreting the results

| Firm Type | Active Terms | Net Effect | Interpretation |
| --- | --- | --- | --- |
| **Pass P&T, Fail Price** | $\beta_1$ | **`r round(pt_stage_coef, 2)`** | Quality without price competitiveness is punished. |
| **Fail P&T, Pass Price** | $\beta_2$ | **`r round(price_stage_coef, 2)`** | Being cheap isn't enough if clinical value is low. |
| **Pass Both** | $\beta_3$ | **`r round(in_pdl_coef, 2)`** | Satisfying both constraints raises innovation incentives. |