---
title: "MPhil Thesis: HT Week 3"
author: "Mohammad-Shah Karim"
date: today

bibliography: ../references/initial_ideas.bib

execute:
  dir: project

format:
  revealjs:
    css: styles.css
    toc: true
    toc-depth: 1
    slideNumber: true
    auto: true
    scrollable: true
    embed-resources: true
    self-contained-math: true
---

# An Introduction to the ACA 

## What is the Affordable Care Act (ACA)?

- Most comprehensive US health reform since Medicare and Medicaid.
- Expanded access to health insurance and reduced financial barriers to care.
- Intended to improve coverage stability for low-income individuals.

Medicaid expansion under ACA is a **demand shock**

## What changed in Medicaid?

| Feature | Before ACA | After ACA |
|---|---|---|
| Basis for eligibility | Low income and categorical requirement (pregnant, disabled, parent etc.) | Income-based |
| Income threshold | Varied by state, often below 100% FPL | ~138% FPL |
| Coverage for low-income adults without children | Generally not eligible | Eligible if income ≤138% FPL |
| State variation | Significant variation | Reduced variation due to optional expansion |
| Benefit structure | Medicaid covered prescription drugs but rules differed by state | Coverage aligned with essential health benefit standards |

## How was the policy rolled out {#policy-rollout}

- **2010:** ACA enacted
- **2012:** Medicaid expansion becomes state-optional
- **2014:** States begin to adopt expansion

```{r}
#| echo: false
#| results: asis

library(here)
library(htmltools)

includeHTML(here("figures", "rollout_map.html"))
```

Data sourced from: [KFF](https://www.kff.org/affordable-care-act/state-indicator/state-activity-around-expanding-medicaid-under-the-affordable-care-act/?currentTimeframe=0&sortModel=%7B%22colId%22:%22Location%22,%22sort%22:%22asc%22%7D)

## What do we know about the Medicaid expansion & pharmaceutical innovation

@NBERw28755 poses the following questions

1. Do the ACA-induced market size expansion for pharmaceuticals cause an increase in pharmaceutical innovation?

2. Is there any evidence that firms shifted R&D effort toward conditions more affected by Medicaid expansion?

Findings: 

1. No clear change in clinical trial activity
    - For conditions above the median Medicaid market share (ex-ante demand)
    - For top quartile of conditions which faced coverage increases (ex-post demand)

2. Overall: innovation outcomes did not respond to increases in Medicaid covered market size

Unexplored areas:

- We still don’t know how firms respond
- Does Medicaid expansion change where firms innovate?
- Do price regulation weaken market-size incentives?

## Medicaid and Prescription Drugs

**Rebates:**  

  - Brand drugs: ≥ 23.1 % rebate on average manufacturer price or best price 
  - Generics: ≥ 13 % of AMP  
  - Inflation rebate: added if prices rise faster than inflation  

**Preferred Drug Lists:**

- States designate PDLs with negotiated rebates, with some firms adopting a uniform PDL to increase bargaining power.
- Some states join multi-state alliances to collectively negotiate deeper rebates.


# Empirical Analysis

```{r}
library(here)

#exploration/HT_week1_RQ1.r


# Load necessary libraries
library(tidyverse)
library(lubridate)
library(readr)
library(stringr)
library(ggplot2)
library(fixest)
library(knitr)
library(modelsummary)
library(kableExtra)
library(fixest)
library(TwoWayFEWeights)
library(DIDmultiplegtDYN)
library(did)
library(here)


# Load FDA approval data

fda_approvals <- read_csv(
    here(
        "processed_data/fda_approvals/fda_sdud_pdl_merged.csv"
        )
) |>
rename(
    approvals = n_approvals,
    firm = sponsor_name
    )

# Load patent data

patents <- read_csv(
    here(
        "processed_data/patents/patents_sdud_pdl_merged.csv"
    )
) |>
    rename(
        atc_level1 = ATC1,
        patents = num_patents)

# --- Build ex-ante demand --- #

# FDA

fda_ex_ante_stats <- fda_approvals |>
    filter(approval_year <= 2013 & approval_year >= 2010) |>
    distinct(atc_level1, approval_year, approval_quarter, prescriptions)

fda_ex_ante_total_prescriptions <- fda_ex_ante_stats |>
    summarise(
        total_prescriptions = sum(prescriptions, na.rm = TRUE),
        .groups = "drop"
    ) |>
    pull()

fda_ex_ante_demand <- fda_ex_ante_stats |>
    group_by(atc_level1) |>
    summarise(
        d_ex_ante = sum(prescriptions, na.rm = TRUE) / fda_ex_ante_total_prescriptions,
        .groups = "drop"
    )

# Patents

patents_ex_ante_stats <- patents |>
    filter(patent_year <= 2013 & patent_year >= 2010) |>
    distinct(atc_level1, patent_year, patent_quarter, prescriptions)

patents_ex_ante_total_prescriptions <- patents_ex_ante_stats |>
    summarise(
        total_prescriptions = sum(prescriptions, na.rm = TRUE),
        .groups = "drop"
    ) |>
    pull()

patents_ex_ante_demand <- patents_ex_ante_stats |>
    group_by(atc_level1) |>
    summarise(
        d_ex_ante = sum(prescriptions, na.rm = TRUE) / patents_ex_ante_total_prescriptions,
        .groups = "drop"
    )


# --- Build ex-post demand --- #

# FDA

fda_ex_post_stats <- fda_approvals |>
    filter(approval_year %in% 2014:2015) |>
    distinct(atc_level1, approval_year, approval_quarter, prescriptions)

fda_ex_post_total_prescriptions <- fda_ex_post_stats |>
    summarise(
        total_prescriptions = sum(prescriptions, na.rm = TRUE),
        .groups = "drop"
    ) |>
    pull()

fda_ex_post_demand <- fda_ex_post_stats |>
    group_by(atc_level1) |>
    summarise(
        d_ex_post = sum(prescriptions, na.rm = TRUE) / fda_ex_post_total_prescriptions,
        .groups = "drop"
    )

# Patents

patents_ex_post_stats <- patents |>
    filter(patent_year %in% 2014:2015) |>
    distinct(atc_level1, patent_year, patent_quarter, prescriptions)

patents_ex_post_total_prescriptions <- patents_ex_post_stats |>
    summarise(
        total_prescriptions = sum(prescriptions, na.rm = TRUE),
        .groups = "drop"
    ) |>
    pull()

patents_ex_post_demand <- patents_ex_post_stats |>
    group_by(atc_level1) |>
    summarise(
        d_ex_post = sum(prescriptions, na.rm = TRUE) / patents_ex_post_total_prescriptions,
        .groups = "drop"
    )


# --- Merge demand measures --- #

fda_approvals <- fda_approvals |>
  left_join(fda_ex_ante_demand, by = "atc_level1") |>
  left_join(fda_ex_post_demand, by = "atc_level1") |>
  filter(approval_year >= 2010)

patents <- patents |>
  left_join(patents_ex_ante_demand, by = "atc_level1") |>
  left_join(patents_ex_post_demand, by = "atc_level1") |>
  filter(patent_year >= 2010)


# --- Run regressions for FDA data --- #

fda_ex_ante_data <- fda_approvals |>
  select(-d_ex_post) |>
  rename(medicaid_market_share = d_ex_ante)

fda_ex_post_data <- fda_approvals |>
  select(-d_ex_ante) |>
  rename(medicaid_market_share = d_ex_post)

patents_ex_ante_data <- patents |>
  select(-d_ex_post) |>
  rename(medicaid_market_share = d_ex_ante)

patents_ex_post_data <- patents |>
  select(-d_ex_ante) |>
  rename(medicaid_market_share = d_ex_post)


```

## Research Question 1: Overall innovation response to Market size

$$
\begin{aligned}
Y_{f,c,t}
= \alpha_{f, c} + \delta_t
+ \sum_{k \neq 2013}^{2025} \beta_k \left[ \mathbf{1}\{t = k\} \times \text{MedicaidShare}_c \right]
+ \varepsilon_{f,c,t}
\end{aligned}
$$

## Research Question 2: Market Size and Pricing Pressure

$$
\begin{aligned}
Y_{f,c,t}
&= \alpha_{f,c} + \delta_t + \sum_{l \neq 2013}^{2025} \beta_{l}
\left( \text{Expansion}_{l} \times \text{MedicaidShare}_c \right) \\
&\quad + \sum_{j \neq 2013}^{2025} \gamma_{\ell}
\left( \text{Expansion}_{j} \times \text{PDL_Exposure}_{f,c} \right) + \varepsilon_{f,c,t}
\end{aligned}
$$

Where:

$$
\text{PDL_Exposure}_{f,c}
=
\frac{
\text{Number of drugs on PDL}_{f,c}
}{
\text{Number of drugs}_{f,c}
}
$$

- $\beta$ measures market size effects
- $\gamma$ measures the effect of firms more exposed to the PDL pricing pressure

## Research Question 3: Innovation portfolio

$$
\begin{aligned}
\begin{aligned}
Y_{f,c,t}
&= \alpha_f + \lambda_c + \delta_t \\
&\quad + \sum_{c \in \mathcal{C}}
\beta_c
\Big(
\mathbb{1}\{t \geq 2014\}
\times \text{MedicaidShare}_c
\times \mathbb{1}\{\text{ATC} = c\}
\Big)
+ \varepsilon_{f,c,t}.
\end{aligned}
\end{aligned}
$$

# Results

## The puzzle

- Therapeutic classes with greater Medicaid exposure experience a decline in patenting after expansion, even after controlling for PDL exposure

- Market size increased
- Access frictions controlled for
- Innovation falls ?

```{r}


patents_ex_ante_year <- patents_ex_ante_data |>
  group_by(firm, atc_level1, patent_year) |>
  summarise(
    patents = sum(patents, na.rm = TRUE),
    medicaid_market_share = first(medicaid_market_share),
    in_pdl = max(in_pdl, na.rm = TRUE),
    .groups = "drop"
  )


patents_ex_post_year <- patents_ex_post_data |>
  group_by(firm, atc_level1, patent_year) |>
  summarise(
    patents = sum(patents, na.rm = TRUE),
    medicaid_market_share = first(medicaid_market_share),
    in_pdl = max(in_pdl, na.rm = TRUE),
    .groups = "drop"
  )



patent_rq2_feols_exante <- feols(
  patents ~ i(patent_year, medicaid_market_share, ref = 2013)
            + i(patent_year, in_pdl, ref = 2013) | 
              firm^atc_level1 + patent_year,
              data = patents_ex_ante_year,
              cluster = ~ firm
)

iplot(
  patent_rq2_feols_exante,
  i.select = 1,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "Event study: Medicaid market size (ex ante)"
)

iplot(
  patent_rq2_feols_exante,
  i.select = 2,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "Event study: PDL exposure (ex ante)"
)

rm(patent_rq2_feols_exante)

patent_rq2_feols_expost <- feols(
  patents ~ i(patent_year, medicaid_market_share, ref = 2013)
            + i(patent_year, in_pdl, ref = 2013) |  
              firm^atc_level1 + patent_year,
              data = patents_ex_post_year,
              cluster = ~ firm
)

iplot(
  patent_rq2_feols_expost,
  i.select = 1,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "Event study: Medicaid market size (ex post)"
)

iplot(
  patent_rq2_feols_expost,
  i.select = 2,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "Event study: PDL exposure (ex post)"
)

rm(patent_rq2_feols_expost)

```

# Theoretical Framework: Stochastic Innovation and the Bidding Trap

## 1. Production Technology

**Innovation.** Firms choose effort $e_i$ to generate clinical quality $a_i$, subject to idiosyncratic risk $\epsilon_i$:

$$
a_i = \ln(e_i) + \epsilon_i, 
\qquad \epsilon_i \sim N(0,\sigma^2).
$$

**Cost of effort.** Effort is costly and convex:

$$
C(e_i) = \tfrac{1}{2} k e_i^2.
$$

**Clinical gate (P\&T).** A drug is clinically approved (qualifies for the pricing stage) if it meets the threshold $\bar a$:

$$
Q(e_i) \equiv \Pr\!\big(\ln(e_i) + \epsilon_i \ge \bar a\big).
$$

## 2. Three States of the World

Profits depend on the joint realization of clinical outcomes for you $(i)$ and your rival $(j)$.

**State A: Monopoly (the prize).**

- **Condition:** $i$ passes, $j$ fails.  
- **Probability:** $Q(e_i)[1-Q(e_j)]$.  
- **Outcome:** $i$ wins access to market $M$ at a monopoly net price $p^{M}$.

**State B: Duopoly (the auction trap).**

- **Condition:** both $i$ and $j$ pass.  
- **Probability:** $Q(e_i)Q(e_j)$
- **Mechanism:** first-price sealed-bid procurement auction for preferred status (lowest net price wins).  
- **Result (simplifying assumption):** competition drives bids to cost, so expected profit is approximately zero.

**State C: Failure (baseline).**

- **Condition:** $i$ fails the P\&T gate.  
- **Probability:** $1-Q(e_i)$
- **Outcome:** $i$ does not enter; profit $=0$.

## 3. Maximization Problem

The firm chooses effort $e_i$ to maximize expected profit across the three states:

$$
\max_{e_i}\ \Pi_i
=
\underbrace{Q(e_i)\,[1-Q(e_j)] \cdot M\,(p^{M}-c_i)}_{\text{State A: Monopoly}}
+
\underbrace{Q(e_i)Q(e_j)\cdot 0}_{\text{State B: Auction}}
+
\underbrace{[1-Q(e_i)]\cdot 0}_{\text{State C: Failure}}
-
\underbrace{\tfrac{1}{2}k e_i^2}_{\text{Cost of effort}}.
$$

## 4. Simplified Objective

Under the “rent dissipation in duopoly” assumption, the problem collapses to:

$$
\max_{e_i}\ \Pi_i
=
Q(e_i)\,[1-Q(e_j)] \cdot M\,(p^{M}-c_i)
-
\tfrac{1}{2}k e_i^2.
$$


## 5. Visualising the Game

```{mermaid}
flowchart TD

%% Stage 0: Effort
E0["Stage 0: Firms choose effort<br/>e_i , e_j"]

%% Stage 1: Nature draws costs
N1["Stage 1: Nature draws costs<br/>c_i ~ U[c_low, c_high]<br/>c_j ~ U[c_low, c_high]<br/>(Private information)"]

%% Stage 2: Clinical Gate
C1{"Stage 2: Clinical Gate (P&T)"}

M1["State A: Monopoly<br/>i passes, j fails"]
D1["State B: Duopoly (Auction)<br/>Both pass"]
F1["State C: Failure<br/>i fails"]

%% Monopoly pricing
PM["Uncontested outcome<br/>i gets access at p^M"]
PI_M["Profit:<br/>pi_i = M (p^M - c_i)"]

%% Duopoly auction
PD["First-price sealed-bid auction<br/>Firms bid b_i, b_j<br/>Lowest bid wins preferred access"]

W_i["i wins:<br/>b_i < b_j"]
W_j["j wins:<br/>b_j < b_i"]

PI_Di["If i wins:<br/>Expected pi_i approx 0"]
PI_Dj["If j wins:<br/>pi_i = 0"]

%% Failure payoff
PI_F["pi_i = 0"]

%% Flow connections
E0 --> N1
N1 --> C1

C1 -->|i passes, j fails| M1
C1 -->|both pass| D1
C1 -->|i fails| F1

M1 --> PM
PM --> PI_M

D1 --> PD
PD --> W_i
PD --> W_j
W_i --> PI_Di
W_j --> PI_Dj

F1 --> PI_F
```


# Bibliography

## References