---
title: "MPhil Thesis: HT Week 3"
author: "Mohammad-Shah Karim"
date: today

bibliography: ../references/initial_ideas.bib

execute:
  dir: project

format:
  revealjs:
    css: styles.css
    toc: true
    toc-depth: 1
    slideNumber: true
    auto: true
    scrollable: true
    embed-resources: true
    self-contained-math: true
---

# An Introduction to the ACA 

## What is the Affordable Care Act (ACA)?

- Most comprehensive US health reform since Medicare and Medicaid.
- Expanded access to health insurance and reduced financial barriers to care.
- Intended to improve coverage stability for low-income individuals.

Medicaid expansion under ACA is a **demand shock**

## What changed in Medicaid?

| Feature | Before ACA | After ACA |
|---|---|---|
| Basis for eligibility | Low income and categorical requirement (pregnant, disabled, parent etc.) | Income-based |
| Income threshold | Varied by state, often below 100% FPL | ~138% FPL |
| Coverage for low-income adults without children | Generally not eligible | Eligible if income ≤138% FPL |
| State variation | Significant variation | Reduced variation due to optional expansion |
| Benefit structure | Medicaid covered prescription drugs but rules differed by state | Coverage aligned with essential health benefit standards |

## How was the policy rolled out {#policy-rollout}

- **2010:** ACA enacted
- **2012:** Medicaid expansion becomes state-optional
- **2014:** States begin to adopt expansion

```{r}
#| echo: false
#| results: asis

library(here)
library(htmltools)

includeHTML(here("figures", "rollout_map.html"))
```

Data sourced from: [KFF](https://www.kff.org/affordable-care-act/state-indicator/state-activity-around-expanding-medicaid-under-the-affordable-care-act/?currentTimeframe=0&sortModel=%7B%22colId%22:%22Location%22,%22sort%22:%22asc%22%7D)

## What do we know about the Medicaid expansion & pharmaceutical innovation

@NBERw28755 poses the following questions

1. Do the ACA-induced market size expansion for pharmaceuticals cause an increase in pharmaceutical innovation?

2. Is there any evidence that firms shifted R&D effort toward conditions more affected by Medicaid expansion?

Findings: 

1. No clear change in clinical trial activity
    - For conditions above the median Medicaid market share (ex-ante demand)
    - For top quartile of conditions which faced coverage increases (ex-post demand)

2. Overall: innovation outcomes did not respond to increases in Medicaid covered market size

Unexplored areas:

- We still don’t know how firms respond
- Does Medicaid expansion change where firms innovate?
- Do price regulation weaken market-size incentives?

## Medicaid and Prescription Drugs

**Rebates:**  

  - Brand drugs: ≥ 23.1 % rebate on average manufacturer price or best price 
  - Generics: ≥ 13 % of AMP  
  - Inflation rebate: added if prices rise faster than inflation  

**Preferred Drug Lists:**

- States designate PDLs with negotiated rebates, with some firms adopting a uniform PDL to increase bargaining power.
- Some states join multi-state alliances to collectively negotiate deeper rebates.


# Empirical Analysis

```{r}
library(here)

#exploration/HT_week1_RQ1.r


# Load necessary libraries
library(tidyverse)
library(lubridate)
library(readr)
library(stringr)
library(ggplot2)
library(fixest)
library(knitr)
library(modelsummary)
library(kableExtra)
library(fixest)
library(TwoWayFEWeights)
library(DIDmultiplegtDYN)
library(did)
library(here)


# Load FDA approval data

fda_approvals <- read_csv(
    here(
        "processed_data/fda_approvals/fda_sdud_pdl_merged.csv"
        )
) |>
rename(
    approvals = n_approvals,
    firm = sponsor_name
    )

# Load patent data

patents <- read_csv(
    here(
        "processed_data/patents/patents_sdud_pdl_merged.csv"
    )
) |>
    rename(
        atc_level1 = ATC1,
        patents = num_patents)

# --- Build ex-ante demand --- #

# FDA

fda_ex_ante_stats <- fda_approvals |>
    filter(approval_year <= 2013 & approval_year >= 2010) |>
    distinct(atc_level1, approval_year, approval_quarter, prescriptions)

fda_ex_ante_total_prescriptions <- fda_ex_ante_stats |>
    summarise(
        total_prescriptions = sum(prescriptions, na.rm = TRUE),
        .groups = "drop"
    ) |>
    pull()

fda_ex_ante_demand <- fda_ex_ante_stats |>
    group_by(atc_level1) |>
    summarise(
        d_ex_ante = sum(prescriptions, na.rm = TRUE) / fda_ex_ante_total_prescriptions,
        .groups = "drop"
    )

# Patents

patents_ex_ante_stats <- patents |>
    filter(patent_year <= 2013 & patent_year >= 2010) |>
    distinct(atc_level1, patent_year, patent_quarter, prescriptions)

patents_ex_ante_total_prescriptions <- patents_ex_ante_stats |>
    summarise(
        total_prescriptions = sum(prescriptions, na.rm = TRUE),
        .groups = "drop"
    ) |>
    pull()

patents_ex_ante_demand <- patents_ex_ante_stats |>
    group_by(atc_level1) |>
    summarise(
        d_ex_ante = sum(prescriptions, na.rm = TRUE) / patents_ex_ante_total_prescriptions,
        .groups = "drop"
    )


# --- Build ex-post demand --- #

# FDA

fda_ex_post_stats <- fda_approvals |>
    filter(approval_year %in% 2014:2015) |>
    distinct(atc_level1, approval_year, approval_quarter, prescriptions)

fda_ex_post_total_prescriptions <- fda_ex_post_stats |>
    summarise(
        total_prescriptions = sum(prescriptions, na.rm = TRUE),
        .groups = "drop"
    ) |>
    pull()

fda_ex_post_demand <- fda_ex_post_stats |>
    group_by(atc_level1) |>
    summarise(
        d_ex_post = sum(prescriptions, na.rm = TRUE) / fda_ex_post_total_prescriptions,
        .groups = "drop"
    )

# Patents

patents_ex_post_stats <- patents |>
    filter(patent_year %in% 2014:2015) |>
    distinct(atc_level1, patent_year, patent_quarter, prescriptions)

patents_ex_post_total_prescriptions <- patents_ex_post_stats |>
    summarise(
        total_prescriptions = sum(prescriptions, na.rm = TRUE),
        .groups = "drop"
    ) |>
    pull()

patents_ex_post_demand <- patents_ex_post_stats |>
    group_by(atc_level1) |>
    summarise(
        d_ex_post = sum(prescriptions, na.rm = TRUE) / patents_ex_post_total_prescriptions,
        .groups = "drop"
    )


# --- Merge demand measures --- #

fda_approvals <- fda_approvals |>
  left_join(fda_ex_ante_demand, by = "atc_level1") |>
  left_join(fda_ex_post_demand, by = "atc_level1") |>
  filter(approval_year >= 2010)

patents <- patents |>
  left_join(patents_ex_ante_demand, by = "atc_level1") |>
  left_join(patents_ex_post_demand, by = "atc_level1") |>
  filter(patent_year >= 2010)


# --- Run regressions for FDA data --- #

fda_ex_ante_data <- fda_approvals |>
  select(-d_ex_post) |>
  rename(medicaid_market_share = d_ex_ante)

fda_ex_post_data <- fda_approvals |>
  select(-d_ex_ante) |>
  rename(medicaid_market_share = d_ex_post)

patents_ex_ante_data <- patents |>
  select(-d_ex_post) |>
  rename(medicaid_market_share = d_ex_ante)

patents_ex_post_data <- patents |>
  select(-d_ex_ante) |>
  rename(medicaid_market_share = d_ex_post)


```

## Research Question 1: Overall innovation response to Market size

$$
\begin{aligned}
Y_{f,c,t}
= \alpha_{f, c} + \delta_t
+ \sum_{k \neq 2013}^{2025} \beta_k \left[ \mathbf{1}\{t = k\} \times \text{MedicaidShare}_c \right]
+ \varepsilon_{f,c,t}
\end{aligned}
$$

## Research Question 2: Market Size and Pricing Pressure

$$
\begin{aligned}
Y_{f,c,t}
&= \alpha_{f,c} + \delta_t + \sum_{l \neq 2013}^{2025} \beta_{l}
\left( \text{Expansion}_{l} \times \text{MedicaidShare}_c \right) \\
&\quad + \sum_{j \neq 2013}^{2025} \gamma_{\ell}
\left( \text{Expansion}_{j} \times \text{PDL_Exposure}_{f,c} \right) + \varepsilon_{f,c,t}
\end{aligned}
$$

Where:

$$
\text{PDL_Exposure}_{f,c}
=
\frac{
\text{Number of drugs on PDL}_{f,c}
}{
\text{Number of drugs}_{f,c}
}
$$

- $\beta$ measures market size effects
- $\gamma$ measures the effect of firms more exposed to the PDL pricing pressure

## Research Question 3: Innovation portfolio

$$
\begin{aligned}
\begin{aligned}
Y_{f,c,t}
&= \alpha_f + \lambda_c + \delta_t \\
&\quad + \sum_{c \in \mathcal{C}}
\beta_c
\Big(
\mathbb{1}\{t \geq 2014\}
\times \text{MedicaidShare}_c
\times \mathbb{1}\{\text{ATC} = c\}
\Big)
+ \varepsilon_{f,c,t}.
\end{aligned}
\end{aligned}
$$

# Results

## The puzzle

- Therapeutic classes with greater Medicaid exposure experience a decline in patenting after expansion, even after controlling for PDL exposure

- Market size increased
- Access frictions controlled for
- Innovation falls ?

```{r}


patents_ex_ante_year <- patents_ex_ante_data |>
  group_by(firm, atc_level1, patent_year) |>
  summarise(
    patents = sum(patents, na.rm = TRUE),
    medicaid_market_share = first(medicaid_market_share),
    in_pdl = max(in_pdl, na.rm = TRUE),
    .groups = "drop"
  )


patents_ex_post_year <- patents_ex_post_data |>
  group_by(firm, atc_level1, patent_year) |>
  summarise(
    patents = sum(patents, na.rm = TRUE),
    medicaid_market_share = first(medicaid_market_share),
    in_pdl = max(in_pdl, na.rm = TRUE),
    .groups = "drop"
  )



patent_rq2_feols_exante <- feols(
  patents ~ i(patent_year, medicaid_market_share, ref = 2013)
            + i(patent_year, in_pdl, ref = 2013) | 
              firm^atc_level1 + patent_year,
              data = patents_ex_ante_year,
              cluster = ~ firm
)

iplot(
  patent_rq2_feols_exante,
  i.select = 1,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "Event study: Medicaid market size (ex ante)"
)

iplot(
  patent_rq2_feols_exante,
  i.select = 2,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "Event study: PDL exposure (ex ante)"
)

rm(patent_rq2_feols_exante)

patent_rq2_feols_expost <- feols(
  patents ~ i(patent_year, medicaid_market_share, ref = 2013)
            + i(patent_year, in_pdl, ref = 2013) |  
              firm^atc_level1 + patent_year,
              data = patents_ex_post_year,
              cluster = ~ firm
)

iplot(
  patent_rq2_feols_expost,
  i.select = 1,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "Event study: Medicaid market size (ex post)"
)

iplot(
  patent_rq2_feols_expost,
  i.select = 2,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "Event study: PDL exposure (ex post)"
)

rm(patent_rq2_feols_expost)

```

# Theoretical Framework

## 1. Production Technology: Stochastic Innovation

**The Core Tension:**

Innovation is a function of investment **effort** ($e_i$) and **risk** ($\epsilon_i$).

**The Quality Production Function:**

$$
a_i = \ln(e_i) + \epsilon_i, \quad \epsilon_i \sim N(0, \sigma^2)
$$

- **Effort ($e_i$):** Deterministic investment (R&D spending).
- **Shock ($\epsilon_i$):** Idiosyncratic clinical risk (trial success/failure).
- **Cost:** Convex cost of effort, $C(e_i) = \frac{1}{2} k e_i^2$.

**Key Implication:**

Firms choose effort $e_i$ *ex-ante*, before realizing if the drug is a "clinical success" or a failure.

## 2. Institutional Hurdles: The Two Stages

The firm faces two distinct barriers to entry.

**Stage 1: The Clinical Threshold ($\bar{a}$)**

The P&T Committee sets a minimum safety/efficacy standard.

$$
\text{Pass P\&T if } a_i \ge \bar{a}
$$
* Let $Q(e_i) = \Pr(a_i \ge \bar{a})$ be the probability of clinical success.
* $Q'(e_i) > 0$: Higher effort increases the chance of passing.

**Gate 2: The Economic Threshold (Price)**

Among clinically approved drugs, the state awards the PDL based on price.

$$
\text{Win PDL if } p_i < p_j \quad \text{(conditional on passing Gate 1)}
$$

## 3. The Winning Probability ($\omega_i$)

The probability of winning the market ($\omega_i$) decomposes into two mutually exclusive cases:

$$
\omega_i = \underbrace{Q(e_i)[1 - Q(e_j)]}_{\text{Case A: Monopoly}} + \underbrace{Q(e_i)Q(e_j) \cdot \mathbb{I}(p_i < p_j)}_{\text{Case B: Price Competition}}
$$

**The Firm's Incentive:**

- **Case A (Clinical Monopoly):** You pass, rival fails. You win automatically.
  - *High Margins.*
- **Case B (Price War):** Both pass. You compete on price ($p_i$).
  - *Compressed Margins.*

## 4. Firm Optimisation & The "Trap"

**The Maximisation Problem:**
$$
\max_{e_i} \Pi_i = \underbrace{p_i(M) \cdot M \cdot \omega_i}_{\text{Expected Revenue}} - \underbrace{\frac{1}{2}k e_i^2}_{\text{Certain Cost}}
$$

**The Effect of Expansion ($M \uparrow$):**

Expansion increases the state's bargaining power, forcing prices ($p_i$) down in the **Competitive State (Case B)**.

**The Incentive Trap:**

1.  As $p_i \to 0$ (due to expansion), the value of **Case B** collapses.
2.  The firm is left relying only on **Case A** (Monopoly), which is rare.
3.  **Result:** Expected marginal revenue falls below marginal cost. The firm reduces effort $e_i$.

## 5. Mapping Theory to Empirics

This stochastic model explains the regression coefficients perfectly.

| Theoretical Outcome | Empirical Equivalent | Coefficient | Interpretation |
| --- | --- | --- | --- |
| **Effort Sunk, Price War Lost** | **Stranded Quality** | $\beta_1 < 0$ | You paid $\frac{1}{2}ke^2$ to pass P&T, but lost the price war. **Pure Loss.** |
| **Effort Sunk, Price War Won** | **PDL Winner** | $\beta_3 > 0$ | You paid the cost and secured the revenue ($p_i M$). **Success.** |

**Conclusion:**

The negative $\beta_1$ confirms that firms are discouraged by the **risk of winning the science but losing the economics.**

# Bibliography

## References