---
title: "MPhil Thesis: HT Week 5"
author: "Mohammad-Shah Karim"
date: "19 Feb 2026"

bibliography: ../references/initial_ideas.bib

execute:
  dir: project

format:
  revealjs:
    css: styles.css
    toc: true
    toc-depth: 1
    slideNumber: true
    auto: true
    scrollable: true
    embed-resources: true
    self-contained-math: true

engine: thesis
---

```{r}

library(reticulate)
use_condaenv("thesis", required = TRUE)

```

# An Introduction to the ACA 

## What is the Affordable Care Act (ACA)?

- **Medicaid** is a governement health insurance programme covering medical care and prescription drugs. This was previously restricted to specific low-income groups (e.g., children, pregnant women, and the disabled).
- The **ACA** expanded Medicaid eligibility to virtually *all* adults with incomes up to 138% of the federal poverty level, removing the old categorical restrictions.
- Over **21 million** previously uninsured adults gained health coverage and access to prescription drugs—directly through this expansion. 

This expansion represents a **demand shock**

## What changed in Medicaid?

| Feature | Before ACA | After ACA |
|---|---|---|
| Basis for eligibility | Low income and categorical requirement (pregnant, disabled, parent etc.) | Income-based |
| Income threshold | Varied by state, often below 100% FPL | ~138% FPL |
| Coverage for low-income adults without children | Generally not eligible | Eligible if income ≤138% FPL |
| State variation | Significant variation | Reduced variation due to optional expansion |
| Benefit structure | Medicaid covered prescription drugs but rules differed by state | Coverage aligned with essential health benefit standards |

## How was the policy rolled out {#policy-rollout}

- **2010:** ACA enacted
- **2012:** Medicaid expansion becomes state-optional
- **2014:** States begin to adopt expansion

```{r}
#| echo: false
#| results: asis

library(here)
library(htmltools)

includeHTML(here("figures", "rollout_map.html"))
```

Data sourced from: [KFF](https://www.kff.org/affordable-care-act/state-indicator/state-activity-around-expanding-medicaid-under-the-affordable-care-act/?currentTimeframe=0&sortModel=%7B%22colId%22:%22Location%22,%22sort%22:%22asc%22%7D)

## What do we know about the Medicaid expansion & pharmaceutical innovation

@NBERw28755 poses the following questions

1. Do the ACA-induced market size expansion for pharmaceuticals cause an increase in pharmaceutical innovation?

2. Is there any evidence that firms shifted R&D effort toward conditions more affected by Medicaid expansion?

Findings: 

1. No clear change in clinical trial activity
    - For conditions above the median Medicaid market share (ex-ante demand)
    - For top quartile of conditions which faced coverage increases (ex-post demand)

2. Overall: innovation outcomes did not respond to increases in Medicaid covered market size

Unexplored areas:

- We still don’t know how firms respond
- Does Medicaid expansion change where firms innovate?
- Do price regulation weaken market-size incentives?

## The Preferred Drug List (PDL)

The **PDL** is state-managed list that drives prescribing behavior for Medicaid patients.

- Drugs on the PDL do not require prior authorisation
    - They become the default choice for doctors for a given illness
    - A drug on the list captures the vast majority of the state's market volume.

- In exchange for this guaranteed market volume, firms must compete to offer the state the lowest price net of offered rebates.

A drug becomes preferred when it is clinically acceptable **and** offers the lowest net cost among substitutes.

1. **Clinical Stage**: P&T committee evaluates safety, efficacy and literature on drug.

2. **Price Stage**: States compare net price among clinically equivalent drugs, with price determining which eligible drugs become most preferred.

```{mermaid}
flowchart LR
    Start["Candidate Drug"]
    
    Clinical["Clinical Review<br>(Safety & Efficacy)"]
    Economic["Economic Evaluation<br>(Net Cost vs Substitutes)"]
    
    Start --> Clinical
    Start --> Economic

    Clinical --> Decision{"Composite<br>Decision"}
    Economic --> Decision

    Decision -->|"PASSED BOTH<br>(Clinical + Cost)"| PDL["Preferred Drug<br>(Winner)"]
    Decision -->|"FAILED EITHER<br>(Clinical or Cost)"| NonPDL["Non-Preferred Drug<br>(Counterfactuals)"]
```

# Research Questions

```{r}
library(here)

#exploration/HT_week1_RQ1.r


# Load necessary libraries
library(tidyverse)
library(lubridate)
library(readr)
library(stringr)
library(ggplot2)
library(fixest)
library(knitr)
library(modelsummary)
library(kableExtra)
library(fixest)
library(TwoWayFEWeights)
library(DIDmultiplegtDYN)
library(did)
library(here)


# Load FDA approval data

fda_approvals <- read_csv(
    here(
        "processed_data/fda_approvals/fda_sdud_pdl_merged.csv"
        )
) |>
rename(
    approvals = n_approvals,
    firm = sponsor_name
    )

# Load patent data

patents <- read_csv(
    here(
        "processed_data/patents/patents_sdud_pdl_merged.csv"
    )
) |>
    rename(
        atc_level1 = ATC1,
        patents = num_patents)

# --- Build ex-ante demand --- #

# FDA

fda_ex_ante_stats <- fda_approvals |>
    filter(approval_year <= 2013 & approval_year >= 2010) |>
    distinct(atc_level1, approval_year, approval_quarter, prescriptions)

fda_ex_ante_total_prescriptions <- fda_ex_ante_stats |>
    summarise(
        total_prescriptions = sum(prescriptions, na.rm = TRUE),
        .groups = "drop"
    ) |>
    pull()

fda_ex_ante_demand <- fda_ex_ante_stats |>
    group_by(atc_level1) |>
    summarise(
        d_ex_ante = sum(prescriptions, na.rm = TRUE) / fda_ex_ante_total_prescriptions,
        .groups = "drop"
    )

# Patents

patents_ex_ante_stats <- patents |>
    filter(patent_year <= 2013 & patent_year >= 2010) |>
    distinct(atc_level1, patent_year, patent_quarter, prescriptions)

patents_ex_ante_total_prescriptions <- patents_ex_ante_stats |>
    summarise(
        total_prescriptions = sum(prescriptions, na.rm = TRUE),
        .groups = "drop"
    ) |>
    pull()

patents_ex_ante_demand <- patents_ex_ante_stats |>
    group_by(atc_level1) |>
    summarise(
        d_ex_ante = sum(prescriptions, na.rm = TRUE) / patents_ex_ante_total_prescriptions,
        .groups = "drop"
    )


# --- Build ex-post demand --- #

# FDA

fda_ex_post_stats <- fda_approvals |>
    filter(approval_year %in% 2014:2015) |>
    distinct(atc_level1, approval_year, approval_quarter, prescriptions)

fda_ex_post_total_prescriptions <- fda_ex_post_stats |>
    summarise(
        total_prescriptions = sum(prescriptions, na.rm = TRUE),
        .groups = "drop"
    ) |>
    pull()

fda_ex_post_demand <- fda_ex_post_stats |>
    group_by(atc_level1) |>
    summarise(
        d_ex_post = sum(prescriptions, na.rm = TRUE) / fda_ex_post_total_prescriptions,
        .groups = "drop"
    )

# Patents

patents_ex_post_stats <- patents |>
    filter(patent_year %in% 2014:2015) |>
    distinct(atc_level1, patent_year, patent_quarter, prescriptions)

patents_ex_post_total_prescriptions <- patents_ex_post_stats |>
    summarise(
        total_prescriptions = sum(prescriptions, na.rm = TRUE),
        .groups = "drop"
    ) |>
    pull()

patents_ex_post_demand <- patents_ex_post_stats |>
    group_by(atc_level1) |>
    summarise(
        d_ex_post = sum(prescriptions, na.rm = TRUE) / patents_ex_post_total_prescriptions,
        .groups = "drop"
    )


# --- Merge demand measures --- #

fda_approvals <- fda_approvals |>
  left_join(fda_ex_ante_demand, by = "atc_level1") |>
  left_join(fda_ex_post_demand, by = "atc_level1") |>
  filter(approval_year >= 2010)

patents <- patents |>
  left_join(patents_ex_ante_demand, by = "atc_level1") |>
  left_join(patents_ex_post_demand, by = "atc_level1") |>
  filter(patent_year >= 2010)


# --- Run regressions for FDA data --- #

fda_ex_ante_data <- fda_approvals |>
  select(-d_ex_post) |>
  rename(medicaid_market_share = d_ex_ante)

fda_ex_post_data <- fda_approvals |>
  select(-d_ex_ante) |>
  rename(medicaid_market_share = d_ex_post)

patents_ex_ante_data <- patents |>
  select(-d_ex_post) |>
  rename(medicaid_market_share = d_ex_ante)

patents_ex_post_data <- patents |>
  select(-d_ex_ante) |>
  rename(medicaid_market_share = d_ex_post)


```

## Research Question 1: Overall innovation response to Market size

$$
\begin{aligned}
Y_{f,c,t}
= \alpha_{f, c} + \delta_t
+ \sum_{k \neq 2013}^{2025} \beta_k \left[ \mathbf{1}\{t = k\} \times \text{MedicaidShare}_c \right]
+ \varepsilon_{f,c,t}
\end{aligned}
$$

Where:

- $Y_{f,c,t}$: innovation outcome for firm $f$ in therapeutic class $c$ and year $t$  
  *(patents or FDA approvals)*
- $\text{MedicaidShare}_c$: ex-ante measure of the utilisation share for therapeutic class $c$
- $\alpha_{f,c}, \delta_t$: firm-disease-class and year fixed effects

## Research Question 2: Market Size and Pricing Pressure

$$
\begin{aligned}
Y_{f,c,t}
&= \alpha_{f,c} + \delta_t + \sum_{l \neq 2013}^{2025} \beta_{l}
\left( \text{Expansion}_{l} \times \text{MedicaidShare}_c \right) \\
&\quad + \sum_{j \neq 2013}^{2025} \gamma_{\ell}
\left( \text{Expansion}_{j} \times \text{PDL_Exposure}_{f,c} \right) + \varepsilon_{f,c,t}
\end{aligned}
$$

Where:

$$
\text{PDL_Exposure}_{f,c}
=
\frac{
\text{Number of drugs on PDL}_{f,c}
}{
\text{Number of drugs}_{f,c}
}
$$

- $\beta$ measures market size effects
- $\gamma$ measures the effect of firms more exposed to the PDL pricing pressure for a given disease class

## Research Question 3: Innovation portfolio

$$
\begin{aligned}
\begin{aligned}
Y_{f,c,t}
&= \alpha_f + \lambda_c + \delta_t \\
&\quad + \sum_{c \in C}
\beta_c
\Big(
\mathbb{1}\{t \geq 2014\}
\times \text{MedicaidShare}_c
\times \mathbb{1}\{\text{ATC} = c\}
\Big)
+ \varepsilon_{f,c,t}
\end{aligned}
\end{aligned}
$$

- $\beta_c$ measures how post-ACA innovation in therapeutic class c responds to its Medicaid exposure.

## The Puzzle - Research Question 2

- Therapeutic classes with greater Medicaid exposure experience a decline in patenting after expansion, even after controlling for PDL exposure
- Market size increased
- Innovation falls

```{r}


patents_ex_ante_year <- patents_ex_ante_data |>
  group_by(firm, atc_level1, patent_year) |>
  summarise(
    patents = sum(patents, na.rm = TRUE),
    medicaid_market_share = first(medicaid_market_share),
    in_pdl = max(in_pdl, na.rm = TRUE),
    .groups = "drop"
  )


patents_ex_post_year <- patents_ex_post_data |>
  group_by(firm, atc_level1, patent_year) |>
  summarise(
    patents = sum(patents, na.rm = TRUE),
    medicaid_market_share = first(medicaid_market_share),
    in_pdl = max(in_pdl, na.rm = TRUE),
    .groups = "drop"
  )



patent_rq2_feols_exante <- feols(
  patents ~ i(patent_year, medicaid_market_share, ref = 2013)
            + i(patent_year, in_pdl, ref = 2013) | 
              firm^atc_level1 + patent_year,
              data = patents_ex_ante_year,
              cluster = ~ firm
)

iplot(
  patent_rq2_feols_exante,
  i.select = 1,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "Event study: Medicaid market size (ex ante)"
)

iplot(
  patent_rq2_feols_exante,
  i.select = 2,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "Event study: PDL exposure (ex ante)"
)

rm(patent_rq2_feols_exante)

patent_rq2_feols_expost <- feols(
  patents ~ i(patent_year, medicaid_market_share, ref = 2013)
            + i(patent_year, in_pdl, ref = 2013) |  
              firm^atc_level1 + patent_year,
              data = patents_ex_post_year,
              cluster = ~ firm
)

iplot(
  patent_rq2_feols_expost,
  i.select = 1,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "Event study: Medicaid market size (ex post)"
)

iplot(
  patent_rq2_feols_expost,
  i.select = 2,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "Event study: PDL exposure (ex post)"
)

rm(patent_rq2_feols_expost)

```

# PDL Counterfactual Analysis

## Understanding counterfactuals

There are two situations by which a pharmaceutical firm gets rejected:

1. **Failing the P&T stage**: The drug does not reach standards for efficacy or safety.

2. **Failing the cost stage**: Some other drug exists that is cheaper for Medicaid to purchase for that state.

In both cases, failing only one of these stages indicates that a firm had just missed out on being eligible for the Medicaid PDL.

It's possible that incentives differ based on which stage a firm failed at.

## Modelling the effect of the PDL

$$
\begin{aligned}
Y_{f,c,t}
&= \alpha_{f,c} + \delta_t \\
&\quad + \beta_1 \big( PT\_Pass \times Post \big) \\
&\quad + \beta_2 \big( Price\_Pass \times Post \big) \\
&\quad + \beta_3 \big( \underbrace{PT\_Pass \times Price\_Pass}_{\text{In PDL}}
\times Post \big)
+ \varepsilon_{f,c,t}
\end{aligned}
$$

- Since $In\_PDL \approx \text{Clinical Pass} \times \text{Price Pass}$, the model is effectively a **Triple Difference** specification.
- This isolates the value of passing each stage independently versus passing both.

## Empirical Analysis

```{r}

rm(list = ls())

# Load necessary libraries
library(tidyverse)
library(lubridate)
library(readr)
library(stringr)
library(ggplot2)
library(fixest)
library(knitr)
library(modelsummary)
library(kableExtra)
library(fixest)
library(TwoWayFEWeights)
library(DIDmultiplegtDYN)
library(did)
library(here)

# Load patent data

patents <- read_csv(
    here(
        "processed_data/patents/patents_sdud_pdl_merged.csv"
    )
) |>
    rename(
        atc_level1 = ATC1,
        patents = num_patents) |>
    filter(patent_year >= 2010)

```

```{r}

patents <- patents |>
    mutate(post = ifelse(patent_year >= 2014, 1, 0))


overall_pdl_test <- feols(
  patents ~ post : pt_stage_pass + price_stage_pass : post + in_pdl : post | 
              firm^atc_level1 + patent_year,
              data = patents,
              cluster = ~ firm
)

modelsummary(overall_pdl_test,
             stars = c(`*` = 0.05, `**` = 0.01, `***` = 0.001),
             gof_omit = "R2 Within*|R2 Adj*|RMSE|Std.Errors|IC|Log")

pt_stage_coef <- as.numeric(coefficients(overall_pdl_test)[1])
price_stage_coef <- as.numeric(coefficients(overall_pdl_test)[2])
in_pdl_coef <- as.numeric(coefficients(overall_pdl_test)[3])

```

## Interpreting the results

| Firm Type | Coefficient | Estimated Effect | Interpretation |
| --- | --- | --- | --- |
| Pass P&T <br> Fail Price | $\beta_1$ | **`r round(pt_stage_coef, 2)`** | Innovation declines post-expansion for drugs that meet clinical standards but fail to be cheap enough. |
| Fail P&T <br> Pass Price | $\beta_2$ | **`r round(price_stage_coef, 2)`** | Innovation declines post-expansion for drugs that are price-competitive but do not pass the clinical stage. |
| Pass P&T <br> Pass Price | $\beta_3$ | **`r round(in_pdl_coef, 2)`** | Innovation increases post-expansion for drugs that secure both clinical approval and preferred pricing status. |


# Theoretical Model

## Stage 1 — Innovation Effort

Firms choose R&D effort $e_i$.

Clinical quality is stochastic:

$$
a_i = \ln(e_i) + \epsilon_i, 
\qquad 
\epsilon_i \sim N(0,\sigma^2)
$$

Effort is costly and convex:

$$
C(e_i) = \frac{1}{2} k e_i^2
$$


## Stage 2 — The P&T Stage

A drug qualifies if it meets threshold $\bar a$.

Probability of passing:

$$
Q(e_i) = \Pr(a_i \ge \bar a)
$$

Effort increases the probability of reaching the pricing stage.

## Stage 3 — The Pricing Test

For firm $i$:

| State | Condition | Value |
|-------|----------|-------|
| Monopoly | $i$ passes, $j$ fails | $V_M$ |
| Duopoly | Both pass | $V_D$ |
| Failure | $i$ fails | $0$ |

Effort $e_i$ increases probability of landing in Monopoly or Duopoly.

## State A: Monopoly

If only firm $i$ passes:

- Firm sets monopoly price $p^M$
- Statutory rebate $\tau(M)$
- Market size $M$

$$
V_M = M \big(p^M - c - \tau(M)\big)
$$

Medicaid expansion affects both:

- $M \uparrow$ (volume effect)
- $\tau(M) \uparrow$ (rebate / monopsony effect)

Net effect ambiguous.

## State B: Duopoly (Procurement Auction)

If both firms pass:

- Firms submit sealed bids $b_i$
- Lowest bid wins PDL status
- Costs are private

$$
c_i \sim F(c) \quad \text{on } [\underline c, \bar c]
$$

Profit for firm $i$:

$$
\pi_i =
\begin{cases}
M(b_i - c_i - \tau(M)) & \text{if } b_i \le b_j \\
0 & \text{otherwise}
\end{cases}
$$

## Auction Equilibrium

Under some symmetric Bayesian Nash equilibrium:

$$
b(c) = c + \tau(M) +
\frac{\int_c^{\bar c} (1 - F(u))\,du}{1 - F(c)}
$$

Expected value in duopoly:

$$
V_D = M \int_c^{\bar c} (1 - F(u))\,du
$$

Interpretation:

- Competition drives prices down
- Only a small margin remains due to private cost differenfes
- Duopoly profits are substantially lower relative to monopoly benchmark

## The Firm’s Maximisation Problem

Firm $i$ chooses effort:

$$
\max_{e_i}
\Pi_i =
Q(e_i)[1 - Q(e_j)]V_M
+
Q(e_i)Q(e_j)V_D
-
\frac{1}{2} k e_i^2
$$

First-order condition:

$$
Q'(e_i)\Big[(1-Q(e_j))V_M + Q(e_j)V_D\Big]
=
k e_i
$$

## Explaining the Puzzle

Expansion ($M \uparrow$) changes incentives via:

$$
V_M = M(p^M - c - \tau(M)),
\qquad
V_D = M \int_c^{\bar c}(1-F(u))du
$$

If rebate growth dominates:

$$
\frac{d}{dM}
\Big[(1-Q)V_M + QV_D\Big] < 0
$$

Then:

$$
\frac{d e^*}{dM} < 0
$$

Innovation falls even though market size increases.

## Mapping Theory to Empirical Results

| Theoretical State | Empirical Proxy | Predicted Sign |
|------------------|-----------------|---------------|
| Exclusive market access (Monopoly state) | PDL Status | $\beta_3 > 0$ |
| Competitive price pressure (Duopoly state) | Passed P&T but not PDL | $\beta_1 < 0$ |

- Innovation declines because the competitive state becomes less valuable.
- Passing P&T no longer guarantees meaningful profits.

# Bibliography

## References