---
title: "MPhil Thesis: HT Week 1"
author: "Mohammad-Shah Karim"
date: "26 Jan 2025"

execute:
  dir: project

format:
  revealjs:
    css: styles.css
    toc: true
    toc-depth: 1
    slideNumber: true
    auto: true
    scrollable: true
    embed-resources: true
    self-contained-math: true
---

```{r}
library(here)

source(here("exploration/HT_week1_RQ1.r"))

```

# Recap

- FDA and Patent data sorted
- Regressions ran for two research questions
    - Effect of ACA on innovation overall
    - Effect of ACA on innovation portfolio
- Failure of parallel trends
- No PDL data
- No price restriction data
- No theoretical model

# Tidying issues from Michaelmas

## Failure of parallel trends

- Fixed patent calculatioins
- Parallel trends for all FEOLS specifications
- Violation of parallel trends under NB patent specifications

# The Preferred Drug list

# A quick reintroduction

- Each state has their own PDL
- Some states form a pool to negotiate better prices

The problem:

- Some PDLs are in pdf format e.g. California
- Some PDLs only show the most recent list e.g. New York

Proposed solution:

- Use Texas Drug Vendor data files

## The General PDL Mechanism revisited

A drug becomes preferred when it is clinically acceptable **and** offers the lowest net cost among substitutes.

1. P&T committee evaluates safety, efficacy and literature on drug.

2. States compare net price among clinically equivalent drugs, with price determining which eligible drugs become most preferred.

```{mermaid}
flowchart LR

A["Clinical Review<br/> (P&T Committee)"] -->|Clinically acceptable| B["Economic Evaluation<br/> (Rebate Negotiation)"]
A -->|Clinically inappropriate| X["Ineligible<br/> (Cannot be preferred)"]

B --> C{"Net cost comparison<br/> among clinically equivalent drugs"}

C -->|Lowest net cost| D["Preferred Drug"]
C -->|Higher net cost| E["Non-preferred Drug<br/> (Authorisation required for use)"]
```

States can list multiple preferred drugs when they are not clinically equivalent.

# Is the mechanism the same in Texas?

Yes but no!

- Both stage 1 and stage 2 must be passed.
- But stage 1 and stage 2 are **not** sequential - they occur simultaenously

This dataset now gives us four scenarios that can be used accordingly:

| Pass efficacy / safety test | Pass cost test | Group
| --- | --- | ----------- |
| ❌ | ❌ | N/A | 
| ❌ | ✅ | Control to measure impact of being cheap on innovation |
| ✅ | ❌ | Control to measure impact of having efficacy on innovation |
| ✅ | ✅ | Treated group |

# Research Questions

## RQ1: Overall innovation response to Market size

$$
\begin{aligned}
Y_{c,t,f}
= \alpha_{c, f} + \delta_t
+ \sum_{k \neq 2013}^{2025} \beta_k \left[ \mathbf{1}\{t = k\} \times \text{MedicaidShare}_c \right]
+ \varepsilon_{c,t,f}
\end{aligned}
$$

- Changed disease-FE to disease $\times$ firm FE
- Firms have different innovation volumes so firm FE makes sense.
- Some therapeutic areas have different innovation volumes.

## Pre-discussion

- Violation of parallel trends for negative binomial model patents model

- Conclusion still consistent with last time
  - No effect on FDA approvals (late stage innovation)
  - Reduction in patents

## RQ1: FDA coefficients

```{r}
fda_rq1_models
```

## RQ1: FDA plots

```{r}

iplot(
  fda_rq1_feols_exante,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "FEOLS FDA approvals ex-ante exposure"
  )

iplot(
  fda_rq1_feols_expost,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "FEOLS FDA approvals ex-post exposure"
  )

iplot(
  fda_rq1_nb_exante,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "NB FDA approvals ex-ante exposure"
  )

iplot(
  fda_rq1_nb_expost,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "NB FDA approvals ex-post exposure"
  )
```

## RQ1: Patent coefficients

```{r}
patent_rq1_models
```

## RQ1: Patent plots

```{r}

iplot(
  patent_rq1_feols_exante,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "FEOLS patents ex-ante exposure"
  )

iplot(
  patent_rq1_feols_expost,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "FEOLS patents ex-post exposure"
  )

iplot(
  patent_rq1_nb_exante,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "NB patents ex-ante exposure"
  )

iplot(
  patent_rq1_nb_expost,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "NB patents ex-post exposure"
  )

```

## RQ2: Market Size and Pricing Pressure

$$
\begin{aligned}
Y_{c,t,f}
&= \alpha_{c,f} + \delta_t + \sum_{l \neq 2013}^{2025} \beta_{l}
\left( \text{Expansion}_{l} \times \text{MedicaidShare}_c \right) \\
&\quad + \sum_{j \neq 2013}^{2025} \gamma_{\ell}
\left( \text{Expansion}_{j} \times \text{PDL_Exposure}_{c,f} \right) + \varepsilon_{c,t,f}
\end{aligned}
$$

Where:

$$
\text{PDL_Exposure}_{c,f}
=
\frac{
\text{Number of drugs on PDL}_{c,f}
}{
\text{Number of drugs}_{c,f}
}
$$

- $\beta$ measures market size effects
- $\gamma$ measures the effect of firms more exposed to the PDL pricing pressure

- Measuring price effects is unrealistic and therefore measuring the effect of being on the market itself is worhtwhl

## Results plot for RQ2


## RQ3: Innovation portfolio









# The theoretical model (WIP)

