---
title: "MPhil Thesis: HT Week 1"
author: "Mohammad-Shah Karim"
date: "26 Jan 2025"

execute:
  dir: project

format:
  revealjs:
    css: styles.css
    toc: true
    toc-depth: 1
    slideNumber: true
    auto: true
    scrollable: true
    embed-resources: true
    self-contained-math: true
---

```{r}
library(here)

source(here("exploration/HT_week1_RQ1.r"))

```

# Recap

- FDA and Patent data sorted
- Regressions ran for two research questions
    - Effect of ACA on innovation overall
    - Effect of ACA on innovation portfolio
- Failure of parallel trends
- No PDL data
- No price restriction data
- No theoretical model

# Tidying issues from Michaelmas

## Failure of parallel trends

- Fixed patent calculatioins
- Parallel trends for all FEOLS specifications
- Violation of parallel trends under NB patent specifications

# The Preferred Drug list

# A quick reintroduction

- Each state has their own PDL
- Some states form a pool to negotiate better prices

The problem:

- Some PDLs are in pdf format e.g. California
- Some PDLs only show the most recent list e.g. New York

Proposed solution:

- Use Texas Drug Vendor data files

## The General PDL Mechanism revisited

A drug becomes preferred when it is clinically acceptable **and** offers the lowest net cost among substitutes.

1. P&T committee evaluates safety, efficacy and literature on drug.

2. States compare net price among clinically equivalent drugs, with price determining which eligible drugs become most preferred.

```{mermaid}
flowchart LR

A["Clinical Review<br/> (P&T Committee)"] -->|Clinically acceptable| B["Economic Evaluation<br/> (Rebate Negotiation)"]
A -->|Clinically inappropriate| X["Ineligible<br/> (Cannot be preferred)"]

B --> C{"Net cost comparison<br/> among clinically equivalent drugs"}

C -->|Lowest net cost| D["Preferred Drug"]
C -->|Higher net cost| E["Non-preferred Drug<br/> (Authorisation required for use)"]
```

States can list multiple preferred drugs when they are not clinically equivalent.

# Is the mechanism the same in Texas?

Yes but no!

- Both stage 1 and stage 2 must be passed.
- But stage 1 and stage 2 are **not** sequential - they occur simultaenously

This dataset now gives us four scenarios that can be used accordingly:

| Pass efficacy / safety test | Pass cost test | Group
| --- | --- | ----------- |
| ❌ | ❌ | N/A | 
| ❌ | ✅ | Control to measure impact of being cheap on innovation |
| ✅ | ❌ | Control to measure impact of having efficacy on innovation |
| ✅ | ✅ | Treated group |

# Research Questions

## RQ1: Overall innovation response to Market size

$$
\begin{aligned}
Y_{f,c,t}
= \alpha_{f, c} + \delta_t
+ \sum_{k \neq 2013}^{2025} \beta_k \left[ \mathbf{1}\{t = k\} \times \text{MedicaidShare}_c \right]
+ \varepsilon_{f,c,t}
\end{aligned}
$$

- Changed disease-FE to disease $\times$ firm FE
- Firms have different innovation volumes so firm FE makes sense.
- Some therapeutic areas have different innovation volumes.

## Pre-discussion

- Violation of parallel trends for negative binomial model patents model

- Conclusion still consistent with last time
  - No effect on FDA approvals (late stage innovation)
  - Reduction in patents

## RQ1: FDA coefficients

```{r}
fda_rq1_models
```

## RQ1: FDA plots

```{r}

iplot(
  fda_rq1_feols_exante,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "FEOLS FDA approvals ex-ante exposure"
  )

iplot(
  fda_rq1_feols_expost,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "FEOLS FDA approvals ex-post exposure"
  )

iplot(
  fda_rq1_nb_exante,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "NB FDA approvals ex-ante exposure"
  )

iplot(
  fda_rq1_nb_expost,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "NB FDA approvals ex-post exposure"
  )
```

## RQ1: Patent coefficients

```{r}
patent_rq1_models
```

## RQ1: Patent plots

```{r}

iplot(
  patent_rq1_feols_exante,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "FEOLS patents ex-ante exposure"
  )

iplot(
  patent_rq1_feols_expost,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "FEOLS patents ex-post exposure"
  )

iplot(
  patent_rq1_nb_exante,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "NB patents ex-ante exposure"
  )

iplot(
  patent_rq1_nb_expost,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "NB patents ex-post exposure"
  )

```

## RQ2: Market Size and Pricing Pressure

$$
\begin{aligned}
Y_{f,c,t}
&= \alpha_{f,c} + \delta_t + \sum_{l \neq 2013}^{2025} \beta_{l}
\left( \text{Expansion}_{l} \times \text{MedicaidShare}_c \right) \\
&\quad + \sum_{j \neq 2013}^{2025} \gamma_{\ell}
\left( \text{Expansion}_{j} \times \text{PDL_Exposure}_{f,c} \right) + \varepsilon_{f,c,t}
\end{aligned}
$$

Where:

$$
\text{PDL_Exposure}_{f,c}
=
\frac{
\text{Number of drugs on PDL}_{f,c}
}{
\text{Number of drugs}_{f,c}
}
$$

- $\beta$ measures market size effects
- $\gamma$ measures the effect of firms more exposed to the PDL pricing pressure

- Measuring price effects is unrealistic and therefore measuring the effect of being on the market itself is worhtwhl


```{r}
source(here("exploration/HT_week1_RQ2.r"))
```

## FDA coefficients

```{r}
fda_rq2_models
```

## FDA plots

```{r}

iplot(
  fda_rq2_feols_exante,
  i.select = 1,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "Event study: Medicaid market size (ex ante)"
)

iplot(
  fda_rq2_feols_exante,
  i.select = 2,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "Event study: PDL exposure (ex ante)"
)

iplot(
  fda_rq2_feols_expost,
  i.select = 1,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "Event study: Medicaid market size (ex post)"
)

iplot(
  fda_rq2_feols_expost,
  i.select = 2,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "Event study: PDL exposure (ex post)"
)

```

## Patent coefficients

```{r}
patent_rq2_models
```

## Patent plots

```{r}

iplot(
  patent_rq2_feols_exante,
  i.select = 1,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "Event study: Medicaid market size (ex ante)"
)

iplot(
  patent_rq2_feols_exante,
  i.select = 2,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "Event study: PDL exposure (ex ante)"
)

iplot(
  patent_rq2_feols_expost,
  i.select = 1,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "Event study: Medicaid market size (ex post)"
)

iplot(
  patent_rq2_feols_expost,
  i.select = 2,
  ci.level = 0.95,
  ref.line = 0,
  xlab = "Year",
  ylab = "Coefficient (relative to 2013)",
  main = "Event study: PDL exposure (ex post)"
)

```

## RQ3: Innovation portfolio

**Old issue:**

Share of innovation variable is polluted by averaging changes in share of innovation across all therapeutic areas. However, a reduction in the share of innovation for therapeutic area A by firm X would also mean that the firm faces an increases in other areas. Current DiD specification cannot isolate these effects .


**Fixes**

1. Change variable from share of innovation to a count of innovation - avoids spillovers effect of an increase in share A leading to a decrease in share B.
2. Introduce triple difference to isolate effects of each therapeutic class.


$$
\begin{aligned}
\begin{aligned}
Y_{f,c,t}
&= \alpha_f + \lambda_c + \delta_t \\
&\quad + \sum_{c \in \mathcal{C}}
\beta_c
\Big(
\mathbb{1}\{t \geq 2014\}
\times \text{MedicaidShare}_c
\times \mathbb{1}\{\text{ATC} = c\}
\Big)
+ \varepsilon_{f,c,t}.
\end{aligned}
\end{aligned}
$$

```{r}
source(here("exploration/HT_week1_RQ3.r"))
```

## FDA coefficients

```{r}

fda_rq3_models

```

## Patent coefficients

```{r}

patent_rq3_models

```

# Appendix

## Available Data

A large issue with cleaning the data comes from cleaning the datasrets in such a way they can be mapped to the Medicaid SDUD dataset.

This involves mapping each of the datasets to a therapeutic area, as well as finding a method to match firms since firms aren't even mentioned in the SDUD data.
  - Standard cleaning and fuzzy matching did not work !

| Outcome Data | Stage in innovation pipeline | Processed |
| -------- | ---- | - |
| Patents | early stage | ✅ | 
| Clinical trials | mid stage | ❌ |
| Drug Approvals | end stage | ✅ | 



## Medicaid State Drug Utilisation Data

| Variable | Info |
| -- | ------------|
| NDC | 11 digit code uniquely identifying drugs, their firm and the package sizing. This is split into three variables: **labeler code** (firm);**product code** (drug for said firm); **package size** |
| Drug name | Name of the drug |
| Units reimbursed | Number of units paid for by Medicaid during the Year-Q |
| Number of prescriptions | Count of prescriptions for a given drug in in that state and Year-Q |
| Reimbursement amount | Tells us the total amount paid for a given drug in that state and Year-Q. This is split into three variables: total, medicaid and non-medicaid.|

Most firms have multiple NDC labeler codes so matching on labeler code is insufficient and further cleaning was required.

## Aggregation

Aggregated at broad therapeutic level.
- ATC level 1: main anatonmical groups
- ATC level 2: further refines each level1 cat into a therapeutic subgroup e.g. diabetes, antibacterials, beta-blockers

| ATC Code | Therapeutic Area |
|-|-------------------|
| A | Alimentary tract and metabolism |
| B | Blood and blood-forming organs |
| C | Cardiovascular system |
| D | Dermatologicals |
| G | Genito-urinary system and sex hormones |
| H | Systemic hormonal preparations |
| J | Anti-infectives for systemic use |
| L | Antineoplastic and immunomodulating agents |
| M | Musculo-skeletal system |
| N | Nervous system |
| P | Antiparasitic products, insecticides and repellents |
| R | Respiratory system |
| S | Sensory organs |
| V | Various |

## Collapsing SDUD

After cleaning Medicaid SDUD data it's collapsed to obtain the number of prescriptions and reimbursement amounts.
This is used to build demand proxies for a given therapeutic area.

```r

# Collapse SDUD data to firm-year-quarter-ATC level

SDUD_firm_ATC <- SDUD_firm_mapped |>
  group_by(
    state,
    firm,
    labeler_code,
    year,
    quarter,
    ATC_Level1,
    ATC_Level2
  ) |>
  summarise(
    prescriptions = sum(number_of_prescriptions, na.rm = TRUE),
    medicaid_reimbursed = sum(medicaid_amount_reimbursed, na.rm = TRUE),
    non_medicaid_reimbursed = sum(non_medicaid_amount_reimbursed, na.rm = TRUE),
    .groups = "drop"
  )

# SDUD ATC level 1 aggregate indicators

SDUD_ATC1 <- SDUD_firm_ATC |>
  group_by(year, quarter, ATC_Level1) |>
  summarise(
    ATC1_prescriptions = sum(prescriptions, na.rm = TRUE),
    ATC1_medicaid_reimbursed = sum(medicaid_reimbursed, na.rm = TRUE),
    ATC1_non_medicaid_reimbursed = sum(non_medicaid_reimbursed, na.rm = TRUE),
    .groups = "drop"
  )

# SDUD ATC level 2 aggregate indicators

SDUD_ATC2 <- SDUD_firm_ATC |>
  group_by(year, quarter, ATC_Level2) |>
  summarise(
    ATC2_prescriptions = sum(prescriptions, na.rm = TRUE),
    ATC2_medicaid_reimbursed = sum(medicaid_reimbursed, na.rm = TRUE),
    ATC2_non_medicaid_reimbursed = sum(non_medicaid_reimbursed, na.rm = TRUE),
    .groups = "drop"
  )

```

## Collapsing Patents

```r

pharma_patents_collapsed <- pharma_patents_full |>
  filter(patent_year >= 2005) |>
  group_by(
    ndc_firm_id,
    firm,
    patent_year,
    patent_quarter,
    ATC1
  ) |>
  summarise(
    num_patents = n(),
    .groups = "drop"
  )

```

## Collapsing Drug approvals

```r

fda_approvals_panel <- fda_atc_expanded |>
  group_by(
    sponsor_name,
    labeler_code,
    approval_year,
    approval_quarter,
    atc_level1,
    atc_level2
  ) |>
  summarise(
    n_approvals = n(),
    .groups = "drop"
  )

```
