---
title: "MPhil Thesis: Initial Ideas"
author: "Mohammad-Shah Karim"
date: "12 Nov 2025"

bibliography: ../references/initial_ideas.bib

execute:
  dir: project


format:
  revealjs:
    css: styles.css
    toc: true
    toc-depth: 1
    slideNumber: true
    auto: true
    scrollable: true
    embed-resources: true
    self-contained-math: true
---

# Introduction

## What is the Affordable Care Act (ACA)?

- Most comprehensive US health reform since Medicare and Medicaid.
- Expanded access to health insurance and reduced financial barriers to care.
- Motivated by high uninsured rates and substantial gaps in Medicaid eligibility across states.
- Introduced subsidised insurance marketplaces for individuals without employer or public coverage.
- Imposed consumer protection rules on insurers including bans on pre-existing condition exclusion and mandating coverage of essentials.
- Intended to improve coverage stability for low-income individuals.

## What changed in Medicaid?

**Before the ACA:**

- Eligibility required low income and belonging to a specific category (e.g., pregnant, disabled, parent of dependent child).
- Income limits varied by state and were often below the Federal Poverty Level. (e.g., Texas threshold was 17% of the FPL)
- Most low-income adults without dependent children were not eligible.

**After the ACA:**

- Eligibility based primarily on income.
- Threshold set at approximately 138% of the Federal Poverty Level.


## Medicaid eligibility

| Feature | Before ACA | After ACA |
|---|---|---|
| Basis for eligibility | Low income and categorical requirement (pregnant, disabled, parent etc.) | Income-based |
| Income threshold | Varied by state, often below 100% FPL | ~138% FPL |
| Coverage for low-income adults without children | Generally not eligible | Eligible if income ≤138% FPL |
| State variation | Significant variation | Reduced variation due to optional expansion |
| Benefit structure | Medicaid covered prescription drugs but rules differed by state | Coverage aligned with essential health benefit standards |

## How was the policy rolled out {#policy-rollout}

- **2010:** ACA enacted
- **2012:** Medicaid expansion becomes state-optional
- **2014:** States begin to adopt expansion

```{r}
#| echo: false
#| include: false

library(here)

source(here("scripts", "rollout_map.r"))

```

```{r}
#| echo: false
#| results: asis

library(here)
library(htmltools)

includeHTML(here("figures", "rollout_map.html"))
```

Data sourced from: [KFF](https://www.kff.org/affordable-care-act/state-indicator/state-activity-around-expanding-medicaid-under-the-affordable-care-act/?currentTimeframe=0&sortModel=%7B%22colId%22:%22Location%22,%22sort%22:%22asc%22%7D)

# Literature Review
## What has the current literature investigated?

| Topic | Findings | Papers |
|---|---|---|
| Insurance coverage & disparities | Large coverage gains concentrated in expansion states; racial/ethnic coverage gaps narrowed. | [@courtemanche2020impact; @frean2017premium; @miller2019four] |
| Access to care | Increased affordability; decreased cost barriers; longer waits and delays | [@miller2017health] |
| Health outcomes | Improvements in self-reported health; reduced adult mortality in expansion states. | [@miller2021medicaid; @miller2017health] |
| Financial outcomes | Declines in medical debt and fewer bankruptcies; improved credit scores. | [@brevoort2017medicaid; @HU201899] |
| Labour market | No significant changes in labour market outcomes; disincentives for labour supply; no change in physician hours. | [@duggan2019effects; @dague2017effect; @neprash2021effect] |

## Unexplored area?

Pharmaceutical innovation induced by the Medicaid expansion !

Only one paper investigates this: @NBERw28755

1. Do the ACA-induced market size expansion for pharmaceuticals cause an increase in pharmaceutical innovation?

2. Is there any evidence that firms shifted R&D effort toward conditions more affected by Medicaid expansion?

Findings: 

1. No clear change in clinical trial activity
    - For conditions above the median Medicaid market share (ex-ante demand)
    - For top quartile of conditions which faced coverage increases (ex-post demand)

2. Overall: innovation outcomes did not respond to increases in Medicaid covered market size

## What are the gaps?

| Gap | Issue in Paper | Proposed Solution |
|---|---|---|
| Short time window | Data up to 2016 | Extend analysis through 2025 |
| No analysis of firm behaviour reallocation | Paper focuses on innovation volume, not whether firms shifted R&D across therapeutic areas | Investigate if firms pivot towards profitable drug classes |
| Medicaid rebate system | Does not study interaction of rebate system and innovation incentives | Separate effects of medicaid monopsony power and increased demand on innovation |

## Medicaid and Prescription Drugs

**Rebates:**  

  - Brand drugs: ≥ 23.1 % rebate on average manufacturer price (AMP) or best price 
  - Generics: ≥ 13 % of AMP  
  - Inflation rebate: added if prices rise faster than inflation  

**Preferred Drug Lists:**

- States designate PDLs with negotiated rebates, with some firms adopting a uniform PDL to increase bargaining power.

**Supplemental Rebate Pools:**  

- States join multi-state alliances to collectively negotiate deeper rebates.

## Research Questions:

**Aim:** Examine how insurance expansion and pricing rules affect pharmaceutical innovation to guide the design of effective policy frameworks.

1.	Does the Medicaid expansion under the ACA affect pharmaceutical innovation over the long run?

2.	Do pharmaceutical firms reallocate R&D effort across therapeutic areas following Medicaid expansion?

3.	How do Medicaid pricing rules influence incentives to innovate?

# Models

## Modelling Choices

Potential models:

- DiD
- Event studies
- Synthetic control
- Callaway and Sant'Anna / Abraham and Sun

Possible outcome variables:

- **RQ1:** Effect on innovation quantity
- **RQ2:** Effect on share of innovation devoted to Medicaid-exposed disease

## Modelling count data

Different distributional assumptions can be made to model the number of patents, clinical trials and drug approvals.

**This needs to be investigated thoroughly.**

Potential censoring issues:

- Censored Poisson
- Poisson thinned
- Pogit

Overdispersion:

- Negative binomial model

Review [learn-bayes](https://github.com/fditraglia/learn-bayes)

## RQ1: Response to Market Size

**Event-study:**

$$
\begin{aligned}
Y_{c,t}
= \alpha_c + \delta_t
+ \sum_{k \neq 2013} \beta_k \left[ \mathbf{1}\{t = k\} \times \text{MedicaidShare}_c \right]
+ \varepsilon_{c,t}
\end{aligned}
$$

where:

- $Y_{c,t}$: innovation outcome for therapeutic class $c$ in year $t$  
  *(patents, clinical trials, FDA approvals)*
- $\text{MedicaidShare}_c$: pre-ACA Medicaid utilisation share for class $c$
- $\alpha_c, \delta_t$: disease-class and year fixed effects



## RQ2: Reallocation Across Therapeutic Areas
$$
\begin{aligned}
\text{Share}_{c,t}
&= \alpha_c + \delta_t
+ \theta \left( \text{PostExpansion}_t \times \text{MedicaidShare}_c \right)
+ \varepsilon_{c,t}
\end{aligned}
$$

Where:

- $\text{Share}_{c,t} = \dfrac{Y_{c,t}}{\sum_{c'} Y_{c',t}}$
- This is a measure for the national share of innovation in disease-class $c$ at time $t$

Feedback: 

- What are the incentives to innovate?
- Beyond the reduced form (i.e. profit seeking behaviour), what is it that results in innovation?
- What is it specific about the context that makes innovation profitable?
- Perhaps consider the different mechanisms at play 
- Used to think of innovation as "there is a disease that has a treatment, is there still some incentive to innovate for these disease classes?" 
- As long as we can make it clear. 
- Is the market expanding for new or old products

## RQ3: Market Size and Pricing Pressure
$$
\begin{aligned}
Y_{c,t}
&= \alpha_c + \delta_t + \sum_{\ell = TBD}^{L} \beta_{\ell}
\left( \text{Expansion}_{t-\ell} \times \text{MedicaidShare}_c \right) \\
&\quad + \sum_{\ell = TBD}^{L} \gamma_{\ell}
\left( PriceRestrict_{t-\ell} \times \text{MedicaidShare}_c \right) + \varepsilon_{c,t}
\end{aligned}
$$

Where:

- $Y_{c,t}$: innovation outcome (patents, trials, approvals) for class $c$ in year $t$
- $\text{Expansion}_{t-\ell}$: Medicaid expansion exposure $\ell$ years prior
- $\text{PriceRestrict}_{t - \ell} = f( \text{UniformPDL}_{t-\ell}, \text{RebatePool}_{t-\ell}, \ \text{MMS}_{t-\ell} )$
$$
\begin{aligned}
\beta_\ell: \text{market-size effects}, \qquad
\gamma_\ell: \text{price-pressure effects}
\end{aligned}
$$

Feedback:

- Perhaps change expansion variable with some variable capturing proportion of population affected by MEdicaid between 0 and 1. 0 in 2013 and 1 if all states adopted. 

## Policy evaluation

- Reduced form allows us to identify direction and magnitude of market-size and pricing effects

- But this could paper could provide insight towards the spillover effects on pharmaceutical innovation and assist with optimal policy design

- Introduce a structural model

- Perhaps this is ambitious

Feedback:

- Does the policy maker feel like there's sufficient competition in the market?
- How did the Medicaid expansion affect market concentration?
- If I did end up on a PDL how did it affect concentration for a given disease class?
- How do you get on the list?
    - try to understand how the PDL list is made.
    - Then we can understand how one gets on the list.
- Do we see innovation harmed / increasing for those that get on the list vs one that does not get on the list?
- How can we then design these things smartly so incentives aren't dampened.

# Data

## Data sources: RQ1

**Effect of Medicaid expansion on innovation**

| Required Data | Source |
|---|---|
| Medicaid expansion timing | KFF |
| Medicaid utilisation share by drug class | CMS State Drug Utilisation Data |
| Patents by disease / drug class | USPTO.gov |
| Clinical trials by condition | ClinicalTrials.gov |
| Drug approvals | Drugs@FDA database |

Feedback:

- Try to find firm level data
- We would like to try and link innovation outcome to the firm that is on / or off the list.
- Try to look for data as well on firms / prescriptions that are not on the drug list. Perhaps prescriptions state-wide for non Medicaid users.
- In long-run try to build a framework where I have two players who face this shock and compete on getting on the list.

## Data sources: RQ2

**Do firms reallocate R&D across therapeutic areas?**

| Required Data | Source |
|---|---|
| Medicaid expansion timing | KFF |
| Baseline Medicaid utilisation share by drug class | CMS State Drug Utilisation Data |
| Drug class disease mapping | RxClass |
| Clinical trials by disease area | ClinicalTrials.gov |
| Patent filings by therapeutic focus | USPTO.gov |

## Data sources: RQ3

**How do Medicaid pricing rules influence innovation?**

| Required Data | Source |
|---|---|
| Medicaid expansion timing | KFF |
| Uniform PDL adoption dates | KFF |
| Rebate-pool membership & timing | KFF |
| Medicaid utilisation share by drug class | CMS State Drug Utilisation Data |

# Bibliography

## References

::: {#refs}
:::