---
title: "MPhil Thesis MT Summary"
author: "Mohammad-Shah Karim"
date: "12 Jan 2025"

bibliography: ../references/MT_summary.bib

execute:
  dir: project

format:
  revealjs:
    css: styles.css
    toc: true
    toc-depth: 1
    slideNumber: true
    auto: true
    scrollable: true
    embed-resources: true
    self-contained-math: true
---

# Introduction

## What is the Affordable Care Act (ACA)?

- Most comprehensive US health reform since Medicare and Medicaid.
- Expanded access to health insurance and reduced financial barriers to care.
- Motivated by high uninsured rates and substantial gaps in Medicaid eligibility across states.
- Introduced subsidised insurance marketplaces for individuals without employer or public coverage.
- Imposed consumer protection rules on insurers including bans on pre-existing condition exclusion and mandating coverage of essentials.
- Intended to improve coverage stability for low-income individuals.

## What changed in Medicaid?

**Before the ACA:**

- Eligibility required low income and belonging to a specific category (e.g., pregnant, disabled, parent of dependent child).
- Income limits varied by state and were often below the Federal Poverty Level. (e.g., Texas threshold was 17% of the FPL)
- Most low-income adults without dependent children were not eligible.

**After the ACA:**

- Eligibility based primarily on income.
- Threshold set at approximately 138% of the Federal Poverty Level.


## Medicaid eligibility

| Feature | Before ACA | After ACA |
|---|---|---|
| Basis for eligibility | Low income and categorical requirement (pregnant, disabled, parent etc.) | Income-based |
| Income threshold | Varied by state, often below 100% FPL | ~138% FPL |
| Coverage for low-income adults without children | Generally not eligible | Eligible if income ≤138% FPL |
| State variation | Significant variation | Reduced variation due to optional expansion |
| Benefit structure | Medicaid covered prescription drugs but rules differed by state | Coverage aligned with essential health benefit standards |

## How was the policy rolled out {#policy-rollout}

- **2010:** ACA enacted
- **2012:** Medicaid expansion becomes state-optional
- **2014:** States begin to adopt expansion

```{r}
#| echo: false
#| include: false

library(here)

source(here("scripts", "rollout_map.r"))

```

```{r}
#| echo: false
#| results: asis

library(here)
library(htmltools)

includeHTML(here("figures", "rollout_map.html"))
```

Data sourced from: [KFF](https://www.kff.org/affordable-care-act/state-indicator/state-activity-around-expanding-medicaid-under-the-affordable-care-act/?currentTimeframe=0&sortModel=%7B%22colId%22:%22Location%22,%22sort%22:%22asc%22%7D)

# Literature Review

## What has the current literature investigated?

| Topic | Findings | Papers |
|---|---|---|
| Insurance coverage & disparities | Large coverage gains concentrated in expansion states; racial/ethnic coverage gaps narrowed. | [@courtemanche2020impact; @frean2017premium; @miller2019four] |
| Access to care | Increased affordability; decreased cost barriers; longer waits and delays | [@miller2017health] |
| Health outcomes | Improvements in self-reported health; reduced adult mortality in expansion states. | [@miller2021medicaid; @miller2017health] |
| Financial outcomes | Declines in medical debt and fewer bankruptcies; improved credit scores. | [@brevoort2017medicaid; @HU201899] |
| Labour market | No significant changes in labour market outcomes; disincentives for labour supply; no change in physician hours. | [@duggan2019effects; @dague2017effect; @neprash2021effect] |

## Unexplored area?

Pharmaceutical innovation induced by the Medicaid expansion !

Only one paper investigates this: @NBERw28755

1. Do the ACA-induced market size expansion for pharmaceuticals cause an increase in pharmaceutical innovation?

2. Is there any evidence that firms shifted R&D effort toward conditions more affected by Medicaid expansion?

Findings: 

1. No clear change in clinical trial activity
    - For conditions above the median Medicaid market share (ex-ante demand)
    - For top quartile of conditions which faced coverage increases (ex-post demand)

2. Overall: innovation outcomes did not respond to increases in Medicaid covered market size

## What are the gaps?

| Gap | Issue in Paper | Proposed Solution |
|---|---|---|
| Short time window | Data up to 2016 | Extend analysis through 2025 |
| No analysis of firm behaviour reallocation | Paper focuses on innovation volume, not whether firms shifted R&D across therapeutic areas | Investigate if firms pivot towards profitable drug classes |
| Medicaid rebate system | Does not study interaction of rebate system and innovation incentives | Separate effects of medicaid monopsony power and increased demand on innovation |


## Research Questions:

**Aim:** Examine how insurance expansion and pricing rules affect pharmaceutical innovation to guide the design of effective policy frameworks.

1.	Does the Medicaid expansion under the ACA affect pharmaceutical innovation over the long run?

2.	Do pharmaceutical firms reallocate R&D effort across therapeutic areas following Medicaid expansion?

3.	How do Medicaid pricing rules influence incentives to innovate?

# Competition within the Pharmaceutical Industry

## Empirical Findings

| Paper | Findings |
|--|-------------|
| @regan2008generic | After initial generic entry, multiple generics enter and branded–generic price gaps widen. Strong post-patent price competition, with substantial price declines. Observes Market segmentation between price-sensitive and less price-sensitive users. |
| @grabowski2007generic | More profitable products attract more generic entrants and have shorter effective exclusivity. Generic competition has become more intense, with many markets seeing rapid erosion of branded share. |
| @kyle2006role | Entry of new drugs into national markets depends on market size and existing competition. Firms are likelier to launch in larger, more profitable markets even when there are incumbents. |
| @OECD2018Pharma | Even during periods of patent exclusivity, on-patent drugs face competition from other products approved for the same indication. Launches of “follow-on” drugs within a therapeutic area create ongoing competitive pressure on prices and market shares. |

## Within Class competition

Drugs within the same therapeutic class face strong competitive pressure, as payers view them as close substitutes and actively reallocate utilisation across on-patent alternatives.


| Mechanism | Source |
|-------------|----|
| Clinically similar drugs constrain each other’s pricing and market share. | @OECD2018Pharma |
| New entrants with small improvements erode incumbent market power. | @OECD2018Pharma |
| Physicians switch within class based on factors such as efficacy, side-effects, guidelines, dosing. | @OECD2018Pharma |

## Pipeline competition

Firms invest in R&D to enter profitable disease areas or to displace incumbents with improved drugs.

| Mechanism | Source |
|-------------|----|
| Entry decisions depend on expected profits and existing competitors | @acemoglu2004market, @dubois2015market |
| Follow-on entry is driven by racing behaviour: multiple firms begin development before the first-in-class drug is approved, aiming to introduce improved variants and capture future market share. | @dimasi2004economics |
| Follow-on drugs intensify competition by weakening the first-mover’s market power and giving payers alternative options to negotiate lower prices. | @OECD2018Pharma |

# Medicaid's Preferred Drug List

## Key facts

**Rebates:**  

  - Brand drugs: ≥ 23.1 % rebate on average manufacturer price (AMP) or best price 
  - Generics: ≥ 13 % of AMP  
  - Inflation rebate: added if prices rise faster than inflation  

**Preferred Drug Lists:**

- States designate PDLs with negotiated rebates, with some firms adopting a uniform PDL to increase bargaining power.

**Supplemental Rebate Pools:**  

- States join multi-state alliances to collectively negotiate deeper rebates.

## Medicaid's Preferred Drug List (PDL)

A drug becomes preferred when it is clinically acceptable **and** offers the lowest net cost among substitutes.

1. P&T committee evaluates safety, efficacy and literature on drug.

2. States compare net price among clinically equivalent drugs, with price determining which eligible drugs become most preferred.

```{mermaid}
flowchart LR

A["Clinical Review<br/> (P&T Committee)"] -->|Clinically acceptable| B["Economic Evaluation<br/> (Rebate Negotiation)"]
A -->|Clinically inappropriate| X["Ineligible<br/> (Cannot be preferred)"]

B --> C{"Net cost comparison<br/> among clinically equivalent drugs"}

C -->|Lowest net cost| D["Preferred Drug"]
C -->|Higher net cost| E["Non-preferred Drug<br/> (Authorisation required for use)"]
```


States can list multiple preferred drugs when they are not clinically equivalent.

## Staying on the PDL

1. Remaining clinically appropriate
  - No concerns relating to being unsafe or outdates
  - No concerns suggesting it's less effective than other drugs

2. Remaining cost competitive
  - The drug must remain as one of the lowest net-cost options among clinically appropriate drugs
  - If a competitor offers a deeper rebate or utilisatioin costs change unfavourably then a drug may be taken off the PDL.

## Do R&D incentives exist for firms on the PDL?

**Preferred status is not guaranteed.**  

Firms can only use R&D to increase the clinical value of their drug:

- Show superior outcomes / fewer side-effects

- Improve adherence or reduced downstream costs
  - e.g. once-daily extended-release instead of twice-daily
- Produce better formulations
  - e.g. pen injector instead of vial administration
- Target specific patient subgroups
  - e.g. better outcomes in patients with renal impairment

However, a firm **cannot** innovate their way into a lower production cost

## R&D incentives for non-PDL firms

Non-preferred status limits utilisation, firms innovate to escape the prior-authorisation barrier and win preferred placement.

Firms with no existing products have strong incentives to enter a profitable therapeutic class.

**Competition methods:**

- Enter with a drug with lower costs of production allowing for a larger rebate than incumbents.
- Develop a drug with better efficacy, or improved dosing.
- Create a drug beneficial for patients not well-served by existing preferred products.


# Data

## Available Data

A large issue with cleaning the data comes from cleaning the datasrets in such a way they can be mapped to the Medicaid SDUD dataset.

This involves mapping each of the datasets to a therapeutic area, as well as finding a method to match firms since firms aren't even mentioned in the SDUD data.
  - Standard cleaning and fuzzy matching did not work !

| Outcome Data | Stage in innovation pipeline | Processed |
| -------- | ---- | - |
| Patents | early stage | ✅ | 
| Clinical trials | mid stage | ❌ |
| Drug Approvals | end stage | ✅ | 



## Medicaid State Drug Utilisation Data

| Variable | Info |
| -- | ------------|
| NDC | 11 digit code uniquely identifying drugs, their firm and the package sizing. This is split into three variables: **labeler code** (firm);**product code** (drug for said firm); **package size** |
| Drug name | Name of the drug |
| Units reimbursed | Number of units paid for by Medicaid during the Year-Q |
| Number of prescriptions | Count of prescriptions for a given drug in in that state and Year-Q |
| Reimbursement amount | Tells us the total amount paid for a given drug in that state and Year-Q. This is split into three variables: total, medicaid and non-medicaid.|

Most firms have multiple NDC labeler codes so matching on labeler code is insufficient and further cleaning was required.

## Aggregation

Aggregated at broad therapeutic level.
- ATC level 1: main anatonmical groups
- ATC level 2: further refines each level1 cat into a therapeutic subgroup e.g. diabetes, antibacterials, beta-blockers

| ATC Code | Therapeutic Area |
|-|-------------------|
| A | Alimentary tract and metabolism |
| B | Blood and blood-forming organs |
| C | Cardiovascular system |
| D | Dermatologicals |
| G | Genito-urinary system and sex hormones |
| H | Systemic hormonal preparations |
| J | Anti-infectives for systemic use |
| L | Antineoplastic and immunomodulating agents |
| M | Musculo-skeletal system |
| N | Nervous system |
| P | Antiparasitic products, insecticides and repellents |
| R | Respiratory system |
| S | Sensory organs |
| V | Various |

## Collapsing SDUD

After cleaning Medicaid SDUD data it's collapsed to obtain the number of prescriptions and reimbursement amounts.
This is used to build demand proxies for a given therapeutic area.

```r

# Collapse SDUD data to firm-year-quarter-ATC level

SDUD_firm_ATC <- SDUD_firm_mapped |>
  group_by(
    state,
    firm,
    labeler_code,
    year,
    quarter,
    ATC_Level1,
    ATC_Level2
  ) |>
  summarise(
    prescriptions = sum(number_of_prescriptions, na.rm = TRUE),
    medicaid_reimbursed = sum(medicaid_amount_reimbursed, na.rm = TRUE),
    non_medicaid_reimbursed = sum(non_medicaid_amount_reimbursed, na.rm = TRUE),
    .groups = "drop"
  )

# SDUD ATC level 1 aggregate indicators

SDUD_ATC1 <- SDUD_firm_ATC |>
  group_by(year, quarter, ATC_Level1) |>
  summarise(
    ATC1_prescriptions = sum(prescriptions, na.rm = TRUE),
    ATC1_medicaid_reimbursed = sum(medicaid_reimbursed, na.rm = TRUE),
    ATC1_non_medicaid_reimbursed = sum(non_medicaid_reimbursed, na.rm = TRUE),
    .groups = "drop"
  )

# SDUD ATC level 2 aggregate indicators

SDUD_ATC2 <- SDUD_firm_ATC |>
  group_by(year, quarter, ATC_Level2) |>
  summarise(
    ATC2_prescriptions = sum(prescriptions, na.rm = TRUE),
    ATC2_medicaid_reimbursed = sum(medicaid_reimbursed, na.rm = TRUE),
    ATC2_non_medicaid_reimbursed = sum(non_medicaid_reimbursed, na.rm = TRUE),
    .groups = "drop"
  )

```

## Collapsing Patents

```r

pharma_patents_collapsed <- pharma_patents_full |>
  filter(patent_year >= 2005) |>
  group_by(
    ndc_firm_id,
    firm,
    patent_year,
    patent_quarter,
    ATC1
  ) |>
  summarise(
    num_patents = n(),
    .groups = "drop"
  )

```

## Collapsing Drug approvals

```r

fda_approvals_panel <- fda_atc_expanded |>
  group_by(
    sponsor_name,
    labeler_code,
    approval_year,
    approval_quarter,
    atc_level1,
    atc_level2
  ) |>
  summarise(
    n_approvals = n(),
    .groups = "drop"
  )

```

# Preliminary Results

```{r}

library(here)

source(here("exploration/MT_week9_modelling.r"))

```

## RQ1: Response to market size

Four models:

- Ex-ante & FEOLS
- Ex-ante & negative binomial
- Ex-post & FEOLS
- Ex-post & negative binomial

## Patents

```{r}

patent_rq1_models

```

## FDA Approvals

```{r}

fda_rq1_models

```

## RQ2: Reallocation Across Therapeutic Areas

```{r}

rq2_models

```

## Interpretation of results

1. Significant reduction in patent applications post ACA
2. No significant effect on FDA approvals post ACA
3. Pharma firms reallocate their share of innovation towards diseases which they don't cover for Medicaid.

**Why did innovation respond negatively to Medicaid expansion?**

1. Medicaid leads to a higher market size but rebates erode profits
    - Firms on the PDL lose incentives to innovate if they gain net-zero profit from drug development.

2. A firm on the PDL has captured the market
    - A dominant firm will not experience a rise in demand from innovation.
    - Investment in innovation to receive the same level of demand would yield less profits compared to a "do-nothing" approach.

Both of these outweigh the benefits of innovating under a larger market size

## Is this consistent with the literature

Results could explain findings in @NBERw28755!

- Their paper sees no significant effect in clinical trials activity post ACA.
- Clinical trials are a late stage outcome and they only use data for two years post treatment
- We only see an effect on patents - the earliest stage outcome in the pharma-innovation pipeline
- We do not observe an effect on the latest stage indicator even after 10 years

This could suggest that innovation is inelastic in the short-run for longer-term indicators.

- Reviewing updated clinical trials data can shed light to this theory

# Next steps

## Theoretical model

Two-firm game involving oligopolistic competition:

1. Expansion of Medicaid as part of ACA is a shock that leads to a jump in market demand
2. Firms innovate to capture market demand
3. However, being on PDL decides whether a firm captures the market or not.
4. How do incentives change and why if a firm captures the market vs if it doesn't.
5. P&T review drugs each year so a drug has the possibility of being removed from the market.

## Steps for Hilary Term

1. Investigate failure of parallel trends to confirm validity of results

2. Clean clinical trials data

3. Obtain Medicaid PDL lists

4. Investigate data on price-rebate intensity

5. Investigate RQ3

6. Build two-player game to explain results


# Bibliography

## References

::: {#refs}
:::