---
title: "MPhil Thesis: MT Week 9"
author: "Mohammad-Shah Karim"
date: "12 Dec 2025"

execute:
  dir: project

format:
  revealjs:
    css: styles.css
    toc: true
    toc-depth: 1
    slideNumber: true
    auto: true
    scrollable: true
    embed-resources: true
    self-contained-math: true
---

# Recap

## Research Questions

1.	Does the Medicaid expansion under the ACA affect pharmaceutical innovation over the long run?

2.	Do pharmaceutical firms reallocate R&D effort across therapeutic areas following Medicaid expansion?

3.	How do Medicaid pricing rules influence incentives to innovate?

## RQ1: Response to Market Size

**Event-study:**

$$
\begin{aligned}
Y_{c,t}
= \alpha_c + \delta_t
+ \sum_{k \neq 2013} \beta_k \left[ \mathbf{1}\{t = k\} \times \text{MedicaidShare}_c \right]
+ \varepsilon_{c,t}
\end{aligned}
$$

where:

- $Y_{c,t}$: innovation outcome for therapeutic class $c$ in year $t$  
  *(patents, clinical trials, FDA approvals)*
- $\text{MedicaidShare}_c$: pre-ACA Medicaid utilisation share for class $c$
- $\alpha_c, \delta_t$: disease-class and year fixed effects



## RQ2: Reallocation Across Therapeutic Areas
$$
\begin{aligned}
\text{Share}_{c,t}
&= \alpha_c + \delta_t
+ \theta \left( \text{PostExpansion}_t \times \text{MedicaidShare}_c \right)
+ \varepsilon_{c,t}
\end{aligned}
$$

Where:

- $\text{Share}_{c,t} = \dfrac{Y_{c,t}}{\sum_{c'} Y_{c',t}}$
- This is a measure for the national share of innovation in disease-class $c$ at time $t$

## RQ3: Market Size and Pricing Pressure
$$
\begin{aligned}
Y_{c,t}
&= \alpha_c + \delta_t + \sum_{\ell = TBD}^{L} \beta_{\ell}
\left( \text{Expansion}_{t-\ell} \times \text{MedicaidShare}_c \right) \\
&\quad + \sum_{\ell = TBD}^{L} \gamma_{\ell}
\left( PriceRestrict_{t-\ell} \times \text{MedicaidShare}_c \right) + \varepsilon_{c,t}
\end{aligned}
$$

Where:

- $Y_{c,t}$: innovation outcome (patents, trials, approvals) for class $c$ in year $t$
- $\text{Expansion}_{t-\ell}$: Medicaid expansion exposure $\ell$ years prior
- $\text{PriceRestrict}_{t - \ell} = f( \text{UniformPDL}_{t-\ell}, \text{RebatePool}_{t-\ell}, \ \text{MMS}_{t-\ell} )$
$$
\begin{aligned}
\beta_\ell: \text{market-size effects}, \qquad
\gamma_\ell: \text{price-pressure effects}
\end{aligned}
$$


## The PDL Mechanism 

A drug becomes preferred when it is clinically acceptable **and** offers the lowest net cost among substitutes.

1. P&T committee evaluates safety, efficacy and literature on drug.

2. States compare net price among clinically equivalent drugs, with price determining which eligible drugs become most preferred.

```{mermaid}
flowchart LR

A["Clinical Review<br/> (P&T Committee)"] -->|Clinically acceptable| B["Economic Evaluation<br/> (Rebate Negotiation)"]
A -->|Clinically inappropriate| X["Ineligible<br/> (Cannot be preferred)"]

B --> C{"Net cost comparison<br/> among clinically equivalent drugs"}

C -->|Lowest net cost| D["Preferred Drug"]
C -->|Higher net cost| E["Non-preferred Drug<br/> (Authorisation required for use)"]
```

# Data

## Available Data

A large issue with cleaning the data comes from cleaning the datasrets in such a way they can be mapped to the Medicaid SDUD dataset.

This involves mapping each of the datasets to a therapeutic area, as well as finding a method to match firms since firms aren't even mentioned in the SDUD data.
  - Standard cleaning and fuzzy matching did not work !

| Outcome Data | Stage in innovation pipeline | Processed |
| -------- | ---- | - |
| Patents | early stage | ✅ | 
| Clinical trials | mid stage | ❌ |
| Drug Approvals | end stage | ✅ | 



## Medicaid State Drug Utilisation Data

| Variable | Info |
| -- | ------------|
| NDC | 11 digit code uniquely identifying drugs, their firm and the package sizing. This is split into three variables: **labeler code** (firm);**product code** (drug for said firm); **package size** |
| Drug name | Name of the drug |
| Units reimbursed | Number of units paid for by Medicaid during the Year-Q |
| Number of prescriptions | Count of prescriptions for a given drug in in that state and Year-Q |
| Reimbursement amount | Tells us the total amount paid for a given drug in that state and Year-Q. This is split into three variables: total, medicaid and non-medicaid.|

Most firms have multiple NDC labeler codes so matching on labeler code is insufficient and further cleaning was required.

## Aggregation

Aggregated at broad therapeutic level.
- ATC level 1: main anatonmical groups
- ATC level 2: further refines each level1 cat into a therapeutic subgroup e.g. diabetes, antibacterials, beta-blockers

| ATC Code | Therapeutic Area |
|-|-------------------|
| A | Alimentary tract and metabolism |
| B | Blood and blood-forming organs |
| C | Cardiovascular system |
| D | Dermatologicals |
| G | Genito-urinary system and sex hormones |
| H | Systemic hormonal preparations |
| J | Anti-infectives for systemic use |
| L | Antineoplastic and immunomodulating agents |
| M | Musculo-skeletal system |
| N | Nervous system |
| P | Antiparasitic products, insecticides and repellents |
| R | Respiratory system |
| S | Sensory organs |
| V | Various |

## Collapsing SDUD

After cleaning Medicaid SDUD data it's collapsed to obtain the number of prescriptions and reimbursement amounts.
This is used to build demand proxies for a given therapeutic area.

```r

# Collapse SDUD data to firm-year-quarter-ATC level

SDUD_firm_ATC <- SDUD_firm_mapped |>
  group_by(
    state,
    firm,
    labeler_code,
    year,
    quarter,
    ATC_Level1,
    ATC_Level2
  ) |>
  summarise(
    prescriptions = sum(number_of_prescriptions, na.rm = TRUE),
    medicaid_reimbursed = sum(medicaid_amount_reimbursed, na.rm = TRUE),
    non_medicaid_reimbursed = sum(non_medicaid_amount_reimbursed, na.rm = TRUE),
    .groups = "drop"
  )

# SDUD ATC level 1 aggregate indicators

SDUD_ATC1 <- SDUD_firm_ATC |>
  group_by(year, quarter, ATC_Level1) |>
  summarise(
    ATC1_prescriptions = sum(prescriptions, na.rm = TRUE),
    ATC1_medicaid_reimbursed = sum(medicaid_reimbursed, na.rm = TRUE),
    ATC1_non_medicaid_reimbursed = sum(non_medicaid_reimbursed, na.rm = TRUE),
    .groups = "drop"
  )

# SDUD ATC level 2 aggregate indicators

SDUD_ATC2 <- SDUD_firm_ATC |>
  group_by(year, quarter, ATC_Level2) |>
  summarise(
    ATC2_prescriptions = sum(prescriptions, na.rm = TRUE),
    ATC2_medicaid_reimbursed = sum(medicaid_reimbursed, na.rm = TRUE),
    ATC2_non_medicaid_reimbursed = sum(non_medicaid_reimbursed, na.rm = TRUE),
    .groups = "drop"
  )

```

## Collapsing Patents

```r

pharma_patents_collapsed <- pharma_patents_full |>
  filter(patent_year >= 2005) |>
  group_by(
    ndc_firm_id,
    firm,
    patent_year,
    patent_quarter,
    ATC1
  ) |>
  summarise(
    num_patents = n(),
    .groups = "drop"
  )

```

## Collapsing Drug approvals

```r

fda_approvals_panel <- fda_atc_expanded |>
  group_by(
    sponsor_name,
    labeler_code,
    approval_year,
    approval_quarter,
    atc_level1,
    atc_level2
  ) |>
  summarise(
    n_approvals = n(),
    .groups = "drop"
  )

```

## Getting on the PDL

Getting on the PDL is a two stage process:

1. safety, efficacy and literature on drug.
2. Being considered cheap enough.

Data from the P&T teams is not available.
-  Cannot observe which firms pass therapeutic stage and which do not.

However - to be used by Medicaid the drug must require authorisation.
- Unlikely to be a perfect substitute as there are likely to be fundamental reasons for said drug not on the list to be chosen instead.
- E.g. the prescriber might insist to use the drug which is not on the list.


# Preliminary Results

```{r}

library(here)

source(here("exploration/MT_week9_modelling.r"))

```

## RQ1: Response to market size

Four models:

- Ex-ante & FEOLS
- Ex-ante & negative binomial
- Ex-post & FEOLS
- Ex-post & negative binomial

## Patents

```{r}

patent_rq1_models

```

## FDA Approvals

```{r}

fda_rq1_models

```

## RQ2: Reallocation Across Therapeutic Areas

```{r}

rq2_models

```

## Steps for next few weeks

1. Clean clinical trials data

2. Obtain Medicaid PDL lists

3. Obtain data for price-rebate intensity

# Feedback

## Romu Comments

- Failure of parallel trends: reduce the number of years used in pre-treatment period.
- No need to use negative binomial for RQ1, FEOLS makes more sense.
- Change outcome from share to count for RQ2 - focus on volume rather than share.
- Change Medicaid variable to identify differential effects across therapeutic area. Some form of triple differencing will be needed for RQ2 to identify this:
  $$
  \text{Present in Medicaid} \times \text{Therapeutic Class} \times \text{Post}
  $$

- Swap ordering of research questions: RQ1, RQ3, RQ2
- Use of drugs that are on SDUD but not PDL should be fine to use as a valid counterfactual given the lack of data.
- Currently using Medicaid prescriptions as demand measure, could consider overall value of demand as well - medicaid amount reimbursed.
- Need to now consider how to build the model: two firm oligopolostic competition game.

## Additional next steps

- Clean up clinical trials data 
- Consider firm-level identification for RQ1
- Build measure of price restriction